### 2025-01-30 article/article.qmd
template

---
title: "Preparing your manuscript"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:
  - name: Michael J Mahoney
    affiliations:
      - name: State University of New York College of Environmental Science and Forestry
        department: Graduate Program in Environmental Science
        address: 1 Forestry Drive
        city: Syracuse
        region: NY
        country: USA
        postal-code: 13210
    orcid: 0000-0003-2402-304X
    email: fake_email@fakeyfake.com
    url: https://mm218.dev
    acknowledgements: Translated template to Quarto.
  - name: Charles Teague
    affiliations:
      - name: Posit, PBC
        address: 250 Northern Ave
        city: Boston
        region: Massachusetts
        country: USA
        postal-code: 02210
    corresponding: true
    orcid: 0000-0001-8715-9476
    email: charles@posit.co
abstract: |
  The abstract (1) states the nature of the investigation and (2) summarizes the important conclusions. The abstract should be suitable for indexing. Your abstract should:

    - Be set as a single paragraph.
    - Be less than 250 words for all journals except GRL, for which the limit is 150 words.
    - Not include table or figure mentions.
    - Avoid reference citations unless dependent on or directly related to another paper (e.g., companion, comment, reply, or commentary on another paper(s)). AGU’s Style Guide discusses formatting citations in abstracts.
    - Define all abbreviations.
plain-language-summary: |
  A Plain Language Summary (PLS) can be an incredibly effective science communication tool. By summarizing your paper in non-technical terms, you can explain your research and its relevance to a much broader audience. A PLS is required for submissions to AGU Advances, G-Cubed, GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets, JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of Geophysics, but optional for other journals.  A PLS should be no longer than 200 words and should be free of jargon, acronyms, equations, and any technical information that would be unknown to people from outside your scientific discipline. Read our tips for creating an effective PLS.
keywords: []
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: bibliography.bib  
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
execute: 
  cache: true
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup

library(metR)
library(ggplot2)
library(data.table)
library(lubridate)
library(here)

global_map <- rnaturalearth::ne_coastline()
models_in_gadi <- c("CMCC-ESM2", "EC-Earth3", "INM-CM4-8", "INM-CM5-0", 
                    "MPI-ESM1-2-HR", "NorESM2-MM", "EC-Earth3-CC", "EC-Earth3-Veg",
                    "EC-Earth3-Veg-LR", "GFDL-CM4")
```

## Introduction

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum hendrerit facilisis velit sit amet malesuada. Phasellus ornare nibh augue, maximus sodales ex tristique vitae. Vivamus non sollicitudin orci, aliquam placerat metus. Maecenas volutpat orci felis, vel finibus urna consectetur sed. Integer in dui ac dui mollis imperdiet. Quisque sed dapibus nibh. Aenean non luctus leo. Phasellus luctus mauris id aliquet dictum. Aliquam fermentum semper massa, vel dignissim nibh dictum et. See @Hubbard2021.

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

In luctus mauris vitae imperdiet luctus. Morbi volutpat ligula ut tortor fermentum, eu ornare felis luctus. Donec semper diam vitae mattis posuere. Suspendisse facilisis purus nisi, sit amet egestas ex tempor ut. Cras tortor nulla, euismod at fermentum vel, dictum vel justo. Aenean commodo interdum diam nec placerat. Nunc vestibulum felis at est tincidunt, at euismod dui vestibulum. Nulla venenatis tortor at auctor iaculis. Donec consectetur neque ut sagittis ornare. Nullam pharetra felis tempor suscipit efficitur. Curabitur nibh ex, euismod at congue hendrerit, egestas id mi. Duis porttitor neque in commodo elementum. Fusce vitae fermentum nisi, euismod viverra augue. Curabitur at mi pretium, accumsan purus nec, tempus turpis.

Donec non semper dui, quis aliquet est. Quisque quis sapien at massa ultricies egestas. Duis consequat ultricies erat, a pulvinar nisl vestibulum id. Sed tristique turpis ligula, et tempor lectus iaculis at. Vivamus commodo sapien ac turpis vestibulum dapibus. Morbi tristique arcu metus, et laoreet nisi varius nec. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Fusce sit amet nisl at mauris suscipit aliquet. Nulla vitae dignissim urna. Suspendisse sit amet arcu vitae magna blandit mattis. Vivamus convallis efficitur pulvinar. Sed cursus elit nulla. Sed porta, arcu a euismod pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex. Nullam consequat ligula sit amet mattis aliquam. Nulla risus urna, ultrices vel ullamcorper id, ornare viverra nunc.

Nunc in lobortis lacus. Duis maximus urna leo, varius sodales arcu interdum nec. Pellentesque imperdiet dolor in leo eleifend dapibus. Ut dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis iaculis risus tellus eget augue. Nullam a viverra est. Cras velit nisi, interdum in lacus at, vehicula mattis elit. Curabitur eu viverra purus. Proin pellentesque, metus vitae congue convallis, lorem metus feugiat mi, sit amet auctor purus ligula bibendum ante. Nam id justo scelerisque, rhoncus lectus in, fermentum libero. Donec tincidunt egestas ex ac eleifend. Cras faucibus ipsum a nunc faucibus fermentum. Integer et maximus lacus. Nam dictum nibh id viverra convallis.

## Data and methods

```{r}
readr::read_rds(here("data/cmip_file_list_scenario.rds")) |> 
  _[nominal_resolution == "100 km" & source_id %in% models_in_gadi, 
    .(source_id, institution_id, nominal_resolution)] |> 
  unique(by = "source_id") |> 
  knitr::kable(col.names = c("Model", "Institution", "Original resolution"),
               caption = "List of models used for the analysis.") |> 
  kableExtra::kable_classic()
```

## Results

## Summary and Discussion

## Acknowledgments

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

## Open research

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

## References {.unnumbered}

:::{#refs}

:::

### 2025-01-30 article/article.qmd
Figure 1, version 1

@@
 library(here)
+library(cowplot)
 
@@
 ```{r}
+#| label: model-table
+
 readr::read_rds(here("data/cmip_file_list_scenario.rds")) |> 
@@
 ## Results
+
+(ref:trend) a) Multimodel change in drops in temperature frequency (%, SSP5-8.5 2018-2100 minus Historical 1980-2000) and zonal mean change over sea b) and land c), each grey line represent the models listed in Table 1 and in black, the multimodel mean.
+
+```{r}
+#| label: trend
+#| fig-width: 10
+#| fig-height: 5
+#| fig.cap: (ref:trend)
+
+data <- purrr::map(Sys.glob(here("data/eke_mean/*hist*mean.nc")), function(f) {
+
+  message(f)
+  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_{scenario}_mean.nc", "{var}_{model}_{scenario}_mean.nc"))
+
+  if (meta[[1]][["var"]] == "eke") {
+    var <- "ua"
+  } else {
+    var <- "tasmax"
+  }
+  file_scenario <- here(paste0("data/eke_mean/", meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585_mean.nc"))
+
+  ReadNetCDF(f, vars = c(hist = var)) |> 
+    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
+    _[, let(model = meta[[1]][["model"]],
+            var = meta[[1]][["var"]],
+            change = ssp585 - hist,
+            lat = round(lat, digits = 8),
+            plev = NULL)] |> 
+    _[]
+
+}) |> rbindlist()
+
+map_trend <- data |> 
+  _[(model %in% models_in_gadi) & is.finite(change)] |>
+  dcast(lat + lon + model ~ var, value.var = "change") |> 
+  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lon, lat)] |> 
+  ggplot(aes(ConvertLongitude(lon), lat)) +
+  geom_contour_fill(aes(z = mean_dt/365*100, fill = after_stat(level)),
+                    breaks = c(-Inf, seq(-3, 3, 0.2), Inf)) +
+  scale_fill_divergent_discretised(guide = guide_colorbar(barheight = 0.5),
+                                   labels = function(x) JumpBy(x, 2, fill = "")) +
+  geom_sf(data = global_map, inherit.aes = FALSE, fill = NA, linewidth = 0.2) +
+  # scale_y_latitude(labels = NULL) +
+  coord_sf(expand = FALSE) +
+  labs(x = NULL, y = NULL, fill = "%") +
+  theme_minimal(base_size = 12) +
+  theme(legend.position = "bottom",
+        legend.frame = ggplot2::element_rect(color = "black", linewidth = 0.4),
+        legend.key.width = grid::unit(1, 'null'),
+        # axis.title.x=element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank(), plot.margin = unit(c(0, 0 , 0 , 0), units = "mm"))
+
+mean_trend <- data |> 
+  _[(model %in% models_in_gadi) & is.finite(change)] |>
+  dcast(lat + lon + model ~ var, value.var = "change") |> 
+  _[, let(land = MaskLand(lon, lat))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
+  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land, model)] |> 
+  ggplot(aes(mean_dt, lat)) +
+  geom_hline(yintercept = 0, color = "grey80") +
+  geom_vline(xintercept = 0, color = "grey80") +
+  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
+  stat_summary(orientation = "y", geom = "line") +
+  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
+  facet_wrap(~land) +
+  tagger::tag_facets(tag_pool = c("b", "c")) +
+  labs(x = "Change (%)") +
+  coord_fixed(ratio = 1/2.8) +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        panel.background = element_rect(fill = "#fbfbfb", color = NA),
+        strip.text.x = element_text(size = 12),
+        plot.margin = unit(c(0, 0 , 1.9, -0.40), units = "cm"),
+        tagger.panel.tag.text = element_text(size = 12),
+         tagger.panel.tag.background = element_rect(fill = "white", color = "white"))
+
+plot_grid(map_trend, mean_trend, labels = c('a)', ''), 
+          label_x = 0, label_y = 0.9,
+          label_fontface = "plain", 
+          label_colour = "white",
+          rel_widths = c(1.75, 1))
+
+```
 

### 2025-01-31 article/article.qmd
Figure 2 - first version

@@
 library(cowplot)
+library(ggside)
 
@@
   geom_sf(data = global_map, inherit.aes = FALSE, fill = NA, linewidth = 0.2) +
-  # scale_y_latitude(labels = NULL) +
   coord_sf(expand = FALSE) +
@@
         legend.key.width = grid::unit(1, 'null'),
-        # axis.title.x=element_blank(),
         axis.text.y = element_blank(),
@@
 
+(ref:serie)
+
+```{r}
+#| label: serie
+#| fig-width: 10
+#| fig-height: 7
+#| fig.cap: (ref:serie)
+
+serie <- purrr::map(Sys.glob(here("data/series/serie_freq_deltap97.5_*_100km.nc")), function(f) {
+
+  meta <- unglue::unglue(basename(f), "serie_freq_{threshold}_{scenario}_100km.nc")
+  ReadNetCDF(f, vars = c(mean_freq = "tasmax")) |> 
+    _[, let(year = year(time),
+            scenario = meta[[1]][["scenario"]])] |> 
+    _[, let(time = NULL)]
+
+}) |> rbindlist() |> 
+  _[, let(scenario = factor(scenario, labels = c("Historical", "SSP5-8.5")))]
+
+model_spread <- data |> 
+  _[, let(land = MaskLand(lon, lat))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
+  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50 - NH",
+                        lat %between% c(38, 40) & land == "Sea", "40 - NH",
+                        lat %between% c(28, 30) & land == "Sea", "30 - NH", 
+                        lat %between% c(48, 50) & land == "Land", "50 - NH",
+                        lat %between% c(38, 40) & land == "Land", "40 - NH",
+                        lat %between% c(28, 30) & land == "Land", "30 - NH",
+                        lat %between% c(-50, -48) & land == "Land", "50 - SH",
+                        lat %between% c(-40, -38) & land == "Land", "40 - SH",
+                        lat %between% c(-30, -28) & land == "Land", "30 - SH",
+                        lat %between% c(-50, -48) & land == "Sea", "50 - SH",
+                        lat %between% c(-40, -38) & land == "Sea", "40 - SH",
+                        lat %between% c(-30, -28) & land == "Sea", "30 - SH"))] |>
+  _[!is.na(serie) & var == "deltap97.5" & model %in% models_in_gadi] |>
+    tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
+  setDT() |> 
+  _[, .(mean_freq = mean(ssp585)), by = .(lat, hemisphere, land, model)] |> 
+  _[, let(scenario = "SSP5-8.5")]
+
+serie_regions <- serie |>
+  _[!(year == 2014 & scenario != "historical")] |>
+  _[, let(land = MaskLand(lon, lat))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
+  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50 - NH",
+                        lat %between% c(38, 40) & land == "Sea", "40 - NH",
+                        lat %between% c(28, 30) & land == "Sea", "30 - NH", 
+                        lat %between% c(48, 50) & land == "Land", "50 - NH",
+                        lat %between% c(38, 40) & land == "Land", "40 - NH",
+                        lat %between% c(28, 30) & land == "Land", "30 - NH",
+                        lat %between% c(-50, -48) & land == "Land", "50 - SH",
+                        lat %between% c(-40, -38) & land == "Land", "40 - SH",
+                        lat %between% c(-30, -28) & land == "Land", "30 - SH",
+                        lat %between% c(-50, -48) & land == "Sea", "50 - SH",
+                        lat %between% c(-40, -38) & land == "Sea", "40 - SH",
+                        lat %between% c(-30, -28) & land == "Sea", "30 - SH"))] |>
+  _[!is.na(serie)] |> 
+  _[, .(mean_freq = mean(mean_freq)), by = .(land, serie, year, scenario)]
+
+serie_regions |> 
+  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
+  setDT() |> 
+  # _[, let(lat = forcats::fct_relevel(lat, "55N", after = 0))] |> 
+  ggplot(aes(year, mean_freq, color = lat, group = interaction(scenario, lat))) +
+  geom_line(alpha = 0.3) +
+  geom_smooth(method = "lm", se = FALSE) +
+  geom_ysideboxplot(data = model_spread, aes(x = lat), orientation = "x") +
+  scale_color_manual(values = c("#004C73", "#CC3311", "#EE7733")) +
+  facet_grid(land~hemisphere, scale = "free") +
+  labs(x = "Year", y = "Frequency (days/year)", color = "Latitude") +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        panel.background = element_rect(fill = "#fbfbfb", color = NA),
+        strip.text = element_text(size = 12),
+        ggside.axis.text.x = element_blank())
+
+```
+
+```{r}
+serie |>
+  _[!(year == 2014 & scenario != "historical") & lat < 0] |>
+  _[, let(land = MaskLand(lon, lat))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
+  _[, let(band_lat = cut_width(lat, 5))] |> 
+  _[, let(mean_freq = mean(mean_freq)), by = .(land, band_lat, year, scenario)] |> 
+  _[, FitLm(mean_freq, year),  by = .(land, band_lat, scenario)] |> 
+  _[term == "year"] |> 
+  ggplot(aes(scenario, estimate)) +
+  geom_point(aes(color = band_lat)) +
+  geom_path(aes(color = band_lat)) +
+  facet_wrap(~land)
+```
+
 ## Summary and Discussion

### 2025-01-31 article/article.qmd
Figure 3 -  firsth version

@@
 
-(ref:serie)
+(ref:serie) Mean frequency of temperature drops for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different latitude bands.
+Left to each main figure the intermodel distribution is shown for the period 2080-2100.
 
@@
 
+(ref:dist) Change in the frequency of drops in temperature (SSP5-8.5 2018-2100 minus Historical 1980-2000) for different percentiles calculates over the historical period (1979-2014).
+
 ```{r}
-serie |>
-  _[!(year == 2014 & scenario != "historical") & lat < 0] |>
-  _[, let(land = MaskLand(lon, lat))] |> 
-  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, let(band_lat = cut_width(lat, 5))] |> 
-  _[, let(mean_freq = mean(mean_freq)), by = .(land, band_lat, year, scenario)] |> 
-  _[, FitLm(mean_freq, year),  by = .(land, band_lat, scenario)] |> 
-  _[term == "year"] |> 
-  ggplot(aes(scenario, estimate)) +
-  geom_point(aes(color = band_lat)) +
-  geom_path(aes(color = band_lat)) +
-  facet_wrap(~land)
+#| label: distribution
+#| fig-width: 10
+#| fig-height: 7
+#| fig.cap: (ref:dist)
+
+file_list <- Sys.glob(here(paste0("data/series/serie_freq_deltap9*", c("historical", "ssp585"), "*")))
+
+dist <- purrr::map(file_list, function(f) {
+  message(f)
+
+  meta <- unglue::unglue(basename(f), "serie_freq_deltap{percentile}_{scenario}_100km.nc")
+
+  if (meta[[1]][["scenario"]] == "historical") {
+
+    ReadNetCDF(f, vars = "tasmax", subset = list(time = c("1980-01-01", "2000-12-31"))) |> 
+      _[, .(mean_freq = mean(tasmax)), by = .(lon, lat)] |> 
+      _[, let(percentile = meta[[1]][["percentile"]],
+              scenario = meta[[1]][["scenario"]])]
+  } else {
+
+    ReadNetCDF(f, vars = "tasmax", subset = list(time = c("2080-01-01", "2100-12-31"))) |> 
+      _[, .(mean_freq = mean(tasmax)), by = .(lon, lat)] |> 
+      _[, let(percentile = meta[[1]][["percentile"]],
+              scenario = meta[[1]][["scenario"]])]
+  }
+
+}) |> rbindlist() |> 
+  _[, let(land = MaskLand(lon, lat),
+          region = fcase(lat %between% c(25, 55), "NH",
+                         lat %between% c(-55, -25), "SH",
+                         default = NA))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))]
+
+dist |> 
+  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50 - NH",
+                        lat %between% c(38, 40) & land == "Sea", "40 - NH",
+                        lat %between% c(28, 30) & land == "Sea", "30 - NH", 
+                        lat %between% c(48, 50) & land == "Land", "50 - NH",
+                        lat %between% c(38, 40) & land == "Land", "40 - NH",
+                        lat %between% c(28, 30) & land == "Land", "30 - NH",
+                        lat %between% c(-50, -48) & land == "Land", "50 - SH",
+                        lat %between% c(-40, -38) & land == "Land", "40 - SH",
+                        lat %between% c(-30, -28) & land == "Land", "30 - SH",
+                        lat %between% c(-50, -48) & land == "Sea", "50 - SH",
+                        lat %between% c(-40, -38) & land == "Sea", "40 - SH",
+                        lat %between% c(-30, -28) & land == "Sea", "30 - SH"))] |>
+  _[!is.na(serie)] |>
+  _[, .(mean_freq = mean(mean_freq)), by = .(land, serie, percentile, scenario)] |> 
+  dcast(land + serie + percentile ~ scenario, value.var = "mean_freq") |> 
+  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
+  setDT() |> 
+  ggplot(aes(percentile, ssp585 / historical)) +
+  geom_col(aes(fill = lat), position = "dodge") +
+  scale_fill_manual(values = c("#004C73", "#CC3311", "#EE7733")) +
+  scale_y_log10() +
+  facet_grid(land~hemisphere) +
+  labs(x = "Percentile", y = "Ratio (SSP5-8.5/Historical)") +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        panel.background = element_rect(fill = "#fbfbfb", color = NA),
+        strip.text = element_text(size = 12))
+
 ```

### 2025-02-03 article/article.qmd
figure 4, maybe?

--- 2025-01-31 article/article.qmd
@@
 library(ggside)
+library(patchwork)
 
@@
 
-(ref:trend) a) Multimodel change in drops in temperature frequency (%, SSP5-8.5 2018-2100 minus Historical 1980-2000) and zonal mean change over sea b) and land c), each grey line represent the models listed in Table 1 and in black, the multimodel mean.
-
 ```{r}
@@
 #| fig-height: 5
-#| fig.cap: (ref:trend)
+#| fig.cap: "a) Multimodel change in drops in temperature frequency (%, SSP5-8.5 2018-2100 minus Historical 1980-2000) and zonal mean change over sea b) and land c), each grey line represent the models listed in Table 1 and in black, the multimodel mean.
+"
 
@@
                     breaks = c(-Inf, seq(-3, 3, 0.2), Inf)) +
-  scale_fill_divergent_discretised(guide = guide_colorbar(barheight = 0.5),
+  scale_fill_divergent_discretised(mid = "#f7f7f7",
+                                   guide = guide_colorbar(barheight = 0.5),
                                    labels = function(x) JumpBy(x, 2, fill = "")) +
@@
 
-(ref:serie)
-
 ```{r}
 #| label: serie
-#| fig-width: 10
-#| fig-height: 7
-#| fig.cap: (ref:serie)
+#| fig-width: 8
+#| fig-height: 5
+#| fig.cap: "Mean frequency of temperature drops for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different latitude bands.
+Left to each main figure the intermodel distribution is shown for the period 2080-2100."
 
@@
   theme_minimal() +
-  theme(legend.position = "bottom",
+  theme(legend.position = "inside",
+        legend.position.inside = c(0.8, 0.9),
+        legend.direction = "horizontal",
+        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
         panel.background = element_rect(fill = "#fbfbfb", color = NA),
@@
 ```{r}
-serie |>
-  _[!(year == 2014 & scenario != "historical") & lat < 0] |>
+#| label: distribution
+#| fig-width: 8
+#| fig-height: 5
+#| fig.cap: "Change in the frequency of drops in temperature (SSP5-8.5 2018-2100 minus Historical 1980-2000) for different percentiles calculates over the historical period (1979-2014)."
+
+file_list <- Sys.glob(here(paste0("data/series/serie_freq_deltap9*", c("historical", "ssp585"), "*")))
+
+dist <- purrr::map(file_list, function(f) {
+  message(f)
+
+  meta <- unglue::unglue(basename(f), "serie_freq_deltap{percentile}_{scenario}_100km.nc")
+
+  if (meta[[1]][["scenario"]] == "historical") {
+
+    ReadNetCDF(f, vars = "tasmax", subset = list(time = c("1980-01-01", "2000-12-31"))) |> 
+      _[, .(mean_freq = mean(tasmax)), by = .(lon, lat)] |> 
+      _[, let(percentile = meta[[1]][["percentile"]],
+              scenario = meta[[1]][["scenario"]])]
+  } else {
+
+    ReadNetCDF(f, vars = "tasmax", subset = list(time = c("2080-01-01", "2100-12-31"))) |> 
+      _[, .(mean_freq = mean(tasmax)), by = .(lon, lat)] |> 
+      _[, let(percentile = meta[[1]][["percentile"]],
+              scenario = meta[[1]][["scenario"]])]
+  }
+
+}) |> rbindlist() |> 
+  _[, let(land = MaskLand(lon, lat),
+          region = fcase(lat %between% c(25, 55), "NH",
+                         lat %between% c(-55, -25), "SH",
+                         default = NA))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))]
+
+dist |> 
+  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50 - NH",
+                        lat %between% c(38, 40) & land == "Sea", "40 - NH",
+                        lat %between% c(28, 30) & land == "Sea", "30 - NH", 
+                        lat %between% c(48, 50) & land == "Land", "50 - NH",
+                        lat %between% c(38, 40) & land == "Land", "40 - NH",
+                        lat %between% c(28, 30) & land == "Land", "30 - NH",
+                        lat %between% c(-50, -48) & land == "Land", "50 - SH",
+                        lat %between% c(-40, -38) & land == "Land", "40 - SH",
+                        lat %between% c(-30, -28) & land == "Land", "30 - SH",
+                        lat %between% c(-50, -48) & land == "Sea", "50 - SH",
+                        lat %between% c(-40, -38) & land == "Sea", "40 - SH",
+                        lat %between% c(-30, -28) & land == "Sea", "30 - SH"))] |>
+  _[!is.na(serie)] |>
+  _[, .(mean_freq = mean(mean_freq)), by = .(land, serie, percentile, scenario)] |> 
+  dcast(land + serie + percentile ~ scenario, value.var = "mean_freq") |> 
+  tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
+  setDT() |> 
+  ggplot(aes(percentile, ssp585 / historical)) +
+  geom_col(aes(fill = lat), position = "dodge") +
+  scale_fill_manual(values = c("#004C73", "#CC3311", "#EE7733")) +
+  scale_y_log10() +
+  facet_grid(land~hemisphere) +
+  labs(x = "Percentile", y = "Ratio (SSP5-8.5/Historical)",
+       fill = "Latitude") +
+  theme_minimal() +
+  theme(legend.position = "inside",
+        legend.position.inside = c(0.8, 0.9),
+        legend.direction = "horizontal",
+        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
+        panel.background = element_rect(fill = "#fbfbfb", color = NA),
+        strip.text = element_text(size = 12))
+
+```
+
+```{r}
+#| label: eke
+#| fig-width: 8
+#| fig-height: 4
+#| fig.cap: "Changes in EKE and drops in temperature frequency for SSP5-8.5 compared with the historical period and each model.
+Values are calculated for northern () and southern () hemisphere and over sea (circles) and land(triangles).
+Each solid line correspond to a linea regresion for each hemisphere.
+"
+
+change <- data |> 
+  _[(model %in% models_in_gadi) & is.finite(change)] |>
+  dcast(lat + lon + model ~ var, value.var = "change") |>
+
+  _[, let(land = MaskLand(lon, lat),
+          region = fcase(lat %between% c(25, 55), "NH",
+                         lat %between% c(-55, -25), "SH",
+                         default = NA))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
+  na.omit() |> 
+  _[, .(mean_eke = mean(eke),
+        mean_deltat = mean(deltap97.5)), by = .(region, land, model)] |> 
+  ggplot(aes(mean_eke, mean_deltat)) +
+  geom_point(aes(color = region, shape = land), size = 2) +
+  geom_smooth(aes(colour = region), method = "lm", se = FALSE) +
+  scale_color_manual(values = c("cyan4", "orange")) +
+  labs(x = "EKE change", y = "Temp drop frequency change",
+       color = NULL, shape = NULL) +
+  theme_minimal() +
+  theme(legend.position = "inside",
+        legend.position.inside = c(0.85, 0.15),
+        legend.direction = "horizontal",
+        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
+        legend.margin = margin(0, 0, 0, 0, unit = "mm"),
+        panel.background = element_rect(fill = "#fbfbfb", color = NA))
+
+deltat_mean <- data |> 
+  _[(model %in% models_in_gadi) & is.finite(change)] |>
+  dcast(lat + lon + model ~ var, value.var = "change") |> 
   _[, let(land = MaskLand(lon, lat))] |> 
   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, let(band_lat = cut_width(lat, 5))] |> 
-  _[, let(mean_freq = mean(mean_freq)), by = .(land, band_lat, year, scenario)] |> 
-  _[, FitLm(mean_freq, year),  by = .(land, band_lat, scenario)] |> 
-  _[term == "year"] |> 
-  ggplot(aes(scenario, estimate)) +
-  geom_point(aes(color = band_lat)) +
-  geom_path(aes(color = band_lat)) +
-  facet_wrap(~land)
+  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land)]
+
+mean_trend_dt <- data |> 
+  _[(model %in% models_in_gadi) & is.finite(change)] |>
+  dcast(lat + lon + model ~ var, value.var = "change") |> 
+  _[, let(land = MaskLand(lon, lat))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
+  _[, .(mean_eke = mean(eke, na.rm = TRUE)), by = .(lat, land, model)] |> 
+  ggplot(aes(mean_eke, lat)) +
+  geom_hline(yintercept = 0, color = "grey80") +
+  geom_vline(xintercept = 0, color = "grey80") +
+  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
+  stat_summary(orientation = "y", geom = "line", aes(color = "EKE")) +
+  geom_path(data = deltat_mean, aes(x = mean_dt, color = "DeltaT")) +
+  scale_color_manual(values = c("DeltaT" = "red4", "EKE" = "black")) +
+  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
+  facet_wrap(~land) +
+  tagger::tag_facets(tag_pool = c("b", "c")) +
+  labs(x = "Change (%)", color = NULL) +
+  coord_fixed(ratio = 1/1.5) +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        panel.background = element_rect(fill = "#fbfbfb", color = NA),
+        strip.text.x = element_text(size = 12),
+        plot.margin = unit(c(0, 0 , 1.9, -0.40), units = "cm"),
+        tagger.panel.tag.text = element_text(size = 12),
+        tagger.panel.tag.background = element_rect(fill = "white", color = "white"))
+
+plot_grid(change, mean_trend_dt, labels = c('a)', ''), 
+          label_x = 0.10, label_y = 0.98,
+          label_fontface = "plain", 
+          label_colour = "black",
+          rel_widths = c(1.75, 1))
 ```

### 2025-02-04 article/article.qmd
draft for the introduction

@@
 author:
-  - name: Michael J Mahoney
+  - name: Paola B Corrales
     affiliations:
-      - name: State University of New York College of Environmental Science and Forestry
-        department: Graduate Program in Environmental Science
-        address: 1 Forestry Drive
-        city: Syracuse
-        region: NY
-        country: USA
-        postal-code: 13210
-    orcid: 0000-0003-2402-304X
+      - name: Monash University
+        department: School of Earth, Atmosphere and Environment
+        region: VIC
+        country: Australia
+    corresponding: true
+    orcid: 0000-0003-1923-9129
     email: fake_email@fakeyfake.com
-    url: https://mm218.dev
-    acknowledgements: Translated template to Quarto.
   - name: Charles Teague
     affiliations:
-      - name: Posit, PBC
-        address: 250 Northern Ave
-        city: Boston
-        region: Massachusetts
-        country: USA
-        postal-code: 02210
-    corresponding: true
+      - name: Monash University
+        department: School of Earth, Atmosphere and Environment
+        region: VIC
+        country: Australia
     orcid: 0000-0001-8715-9476
-    email: charles@posit.co
+    email: martin.singh@monash.edu
 abstract: |
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Vestibulum hendrerit facilisis velit sit amet malesuada.
-Phasellus ornare nibh augue, maximus sodales ex tristique vitae.
-Vivamus non sollicitudin orci, aliquam placerat metus.
-Maecenas volutpat orci felis, vel finibus urna consectetur sed.
-Integer in dui ac dui mollis imperdiet.
-Quisque sed dapibus nibh.
-Aenean non luctus leo.
-Phasellus luctus mauris id aliquet dictum.
-Aliquam fermentum semper massa, vel dignissim nibh dictum et.
-See @Hubbard2021.
+Cold fronts are a important feature of weather in extratropical regions and are in part responsible for the precipitation in these regions [@konstali2024].
+Cold fronts are also linked to extreme events like heavy precipitation [@catto2013] and severe wildfires in southern Australia [@cai2022].
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
+Front identification can be a complex task, to this date, there are several methods that uses different characteristics of the fronts.
+Some of them looks for thermodynamical changes associated with temperature and humidity in different atmospheric levels [@berry2011], others identify changes in the wind field [@simmonds2011].
+These methods relies on the data being in a regular grid (as outputs from numerical models or reanalysis) and usually depends on different thresholds that need to be calibrated for different model resolutions.
+For this study we use the changes from day to day in near surface temperature (or drops in temperature) to estimate the passage of fronts [@reeder2015].
+This very simple method can be use for both data from station observations and gridded data (with varied resolution up to 2.5 degrees [@king2024]).
+Previous studies have shown that this method is suitable for the detection of strong cold fronts in South Eastern Australia [@cai2022, @reader2015], and recently @king2024 showed that this method can successfully identify cold fronts with similar results as the obtained with other methodologies.
+Moreover, @king2024 calculated a climatology of front frequency using global climate models from phase 6 of the Coupled Model Intercomparison Project (CMIP6).
+while the models overestimate the frequency of fronts in polar regions and tend to underestimate the occurrence of large temperature drops over the midlatitude land regions compared with ERA5.
+However, they can represent well the patterns of front frequency equatorward of 60$ ^{\circ}$.
 
-In luctus mauris vitae imperdiet luctus.
-Morbi volutpat ligula ut tortor fermentum, eu ornare felis luctus.
-Donec semper diam vitae mattis posuere.
-Suspendisse facilisis purus nisi, sit amet egestas ex tempor ut.
-Cras tortor nulla, euismod at fermentum vel, dictum vel justo.
-Aenean commodo interdum diam nec placerat.
-Nunc vestibulum felis at est tincidunt, at euismod dui vestibulum.
-Nulla venenatis tortor at auctor iaculis.
-Donec consectetur neque ut sagittis ornare.
-Nullam pharetra felis tempor suscipit efficitur.
-Curabitur nibh ex, euismod at congue hendrerit, egestas id mi.
-Duis porttitor neque in commodo elementum.
-Fusce vitae fermentum nisi, euismod viverra augue.
-Curabitur at mi pretium, accumsan purus nec, tempus turpis.
+As the planet keeps heating up, understanding how the increased global temperature will impact the occurrence of cold fronts is very important.
+A few authors have aproached this question, for example @catto2014 used objective front identification [@berry2011] in 18 CMIP5 models and compared the changes between historical 1980-2005 + RCP8.5 2080-2100.
+They found an overall decrease in front frequency in the Northern Hemisphere, with a poleward shift of the maxima of front frequency and a strong decrease at high latitudes.
+The Southern Hemisphere shows a poleward shift of the frequency maximum.
+On the other hand, @konstali2024, identified fronts using gradient of equivalent potential temperature in 850 hPa for the CESM2-LE model and found an increase in the frequency of fronts related to fronts becoming larger or more elongated in both hemispheres.
 
-Donec non semper dui, quis aliquet est.
-Quisque quis sapien at massa ultricies egestas.
-Duis consequat ultricies erat, a pulvinar nisl vestibulum id.
-Sed tristique turpis ligula, et tempor lectus iaculis at.
-Vivamus commodo sapien ac turpis vestibulum dapibus.
-Morbi tristique arcu metus, et laoreet nisi varius nec.
-Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
-Fusce sit amet nisl at mauris suscipit aliquet.
-Nulla vitae dignissim urna.
-Suspendisse sit amet arcu vitae magna blandit mattis.
-Vivamus convallis efficitur pulvinar.
-Sed cursus elit nulla.
-Sed porta, arcu a euismod pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex.
-Nullam consequat ligula sit amet mattis aliquam.
-Nulla risus urna, ultrices vel ullamcorper id, ornare viverra nunc.
-
-Nunc in lobortis lacus.
-Duis maximus urna leo, varius sodales arcu interdum nec.
-Pellentesque imperdiet dolor in leo eleifend dapibus.
-Ut dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis iaculis risus tellus eget augue.
-Nullam a viverra est.
-Cras velit nisi, interdum in lacus at, vehicula mattis elit.
-Curabitur eu viverra purus.
-Proin pellentesque, metus vitae congue convallis, lorem metus feugiat mi, sit amet auctor purus ligula bibendum ante.
-Nam id justo scelerisque, rhoncus lectus in, fermentum libero.
-Donec tincidunt egestas ex ac eleifend.
-Cras faucibus ipsum a nunc faucibus fermentum.
-Integer et maximus lacus.
-Nam dictum nibh id viverra convallis.
+In this paper we build on the drops in temperature methodology to evaluate the change in frequency of cold fronts in a warmer future compared with the resent past.
+In an effort to understand possible causes to the potential changes, we explored the relationship between changes in front frequency and transient eddy kinetic energy, as a proxy for storm tracks.
+Several studies have shown a poleward shift on the storm tracks, particularly in the Southern Hemisphere [@ogorman2010, @tamarin2017, @chenke2022], and these changes could potentially influence the location and intensity of future cold fronts.
 
@@
 
+Detection of fronts over oceans when using a constant threshold
+Weak fronts are not detected?
+
 ## Acknowledgments

### 2025-02-05 article/article.qmd
fix typos and grammar

@@
   - Key Points are not included in the word count.
-bibliography: bibliography.bib  
+bibliography: cold-fronts.bib  
 citation:
@@
 
-Cold fronts are a important feature of weather in extratropical regions and are in part responsible for the precipitation in these regions [@konstali2024].
-Cold fronts are also linked to extreme events like heavy precipitation [@catto2013] and severe wildfires in southern Australia [@cai2022].
+Cold fronts are a important feature of weather in extratropical regions and are partly responsible for the precipitation in these areas [@konstali2024].
+Cold fronts are also linked to extreme events, such as heavy precipitation [@catto2013] and severe wildfires in southern Australia [@cai2022].
 
-Front identification can be a complex task, to this date, there are several methods that uses different characteristics of the fronts.
-Some of them looks for thermodynamical changes associated with temperature and humidity in different atmospheric levels [@berry2011], others identify changes in the wind field [@simmonds2011].
-These methods relies on the data being in a regular grid (as outputs from numerical models or reanalysis) and usually depends on different thresholds that need to be calibrated for different model resolutions.
-For this study we use the changes from day to day in near surface temperature (or drops in temperature) to estimate the passage of fronts [@reeder2015].
-This very simple method can be use for both data from station observations and gridded data (with varied resolution up to 2.5 degrees [@king2024]).
-Previous studies have shown that this method is suitable for the detection of strong cold fronts in South Eastern Australia [@cai2022, @reader2015], and recently @king2024 showed that this method can successfully identify cold fronts with similar results as the obtained with other methodologies.
-Moreover, @king2024 calculated a climatology of front frequency using global climate models from phase 6 of the Coupled Model Intercomparison Project (CMIP6).
-while the models overestimate the frequency of fronts in polar regions and tend to underestimate the occurrence of large temperature drops over the midlatitude land regions compared with ERA5.
-However, they can represent well the patterns of front frequency equatorward of 60$ ^{\circ}$.
+Identifying fronts can be a complex task.
+To date, several computational methods have been developed that use different characteristics of fronts.
+Some of these methods looks for thermodynamic changes associated with temperature and humidity in different atmospheric levels [@berry2011], others identify changes in the wind field [@simmonds2012].
+These methods rely on the data being in a regular grid (e.g.
+outputs from numerical models or reanalysis) and it results depend on setting different thresholds that need to be calibrated for different model resolutions.
 
-As the planet keeps heating up, understanding how the increased global temperature will impact the occurrence of cold fronts is very important.
-A few authors have aproached this question, for example @catto2014 used objective front identification [@berry2011] in 18 CMIP5 models and compared the changes between historical 1980-2005 + RCP8.5 2080-2100.
-They found an overall decrease in front frequency in the Northern Hemisphere, with a poleward shift of the maxima of front frequency and a strong decrease at high latitudes.
-The Southern Hemisphere shows a poleward shift of the frequency maximum.
-On the other hand, @konstali2024, identified fronts using gradient of equivalent potential temperature in 850 hPa for the CESM2-LE model and found an increase in the frequency of fronts related to fronts becoming larger or more elongated in both hemispheres.
+For this study we use day-to-day changes in near-surface temperature (or drops in temperature) to estimate the passage of fronts [@reeder2015].
+This very simple method can be applied to both data from station observations and gridded data (with varying resolution, up to 2.5 degrees [@king2024]).
+Previous studies have shown that this method is suitable for the detection of strong cold fronts in southeastern Australia [@cai2022, @reeder2015].
+Recently, @king2024 demonstrated that this method can successfully identify cold fronts with results comparable to those obtained with other methodologies.
+Furthermore, @king2024 calculated a climatology of front frequency using global climate models from phase 6 of the Coupled Model Intercomparison Project (CMIP6).
+While the models tend to overestimate front frequency in polar regions and underestimate the occurrence of large temperature drops over midlatitude land regions compared with ERA5, they accurately represent the patterns of front frequency equatorward of 60°.
 
-In this paper we build on the drops in temperature methodology to evaluate the change in frequency of cold fronts in a warmer future compared with the resent past.
-In an effort to understand possible causes to the potential changes, we explored the relationship between changes in front frequency and transient eddy kinetic energy, as a proxy for storm tracks.
-Several studies have shown a poleward shift on the storm tracks, particularly in the Southern Hemisphere [@ogorman2010, @tamarin2017, @chenke2022], and these changes could potentially influence the location and intensity of future cold fronts.
+As global temperatures continue to rise, understanding how this warming will impact the occurrence of cold fronts is very important.
+A few studies have addressed this question, for example @catto2014 used objective front identification [@berry2011] in 18 CMIP5 models and compared the changes between historical simulations between 1980-2005 and the Representative Concentration Pathway 8.5 between 2080-2100.
+They found an overall decrease in front frequency in the Northern Hemisphere, with a poleward shift in the maximum frequency and a significant decrease at high latitudes.
+In contrast, the Southern Hemisphere showed a poleward shift in the frequency maximum.
+On the other hand, @konstali2024, identified fronts in the CESM2-LE model simulations using gradient of equivalent potential temperature in 850 hPa and found an increase in the frequency of fronts that could be related to fronts becoming larger or more elongated in both hemispheres.
+
+In this paper we build on the drops in temperature methodology to evaluate the change in frequency of cold fronts in a warmer future compared with the resent past using a set of models from CMIP6.
+In an effort to understand possible causes to the potential changes, we explored the relationship between changes in front frequency and transient eddy kinetic energy, that is directly related to storm tracks.
+Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@ogorman2010, @tamarin2017, @chemke2022], and these changes could potentially influence the location and intensity of future cold fronts.
 

### 2025-02-10 article/article.qmd
Fix typos and figure small details

@@
     agu-html: default
+    docx: default
 author:
@@
         department: School of Earth, Atmosphere and Environment
-        region: VIC
+        region: Victoria
         country: Australia
@@
     email: fake_email@fakeyfake.com
-  - name: Charles Teague
+  - name: Martin Singh
     affiliations:
@@
 keywords: []
-key-points:
-  - Key Points convey the main points and conclusions of the article. 
-  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
-  - Key Points are not included in the word count.
+keypoints: 
+  - Day-to-day drops in maximun temperature are identified as a proxy for cold fronts.  
+  - SSP5-8.5 from CMIP show a decrease in frequency of drops in temperature on the NH and a poleward shift over the oceans in the SH
+  - storm tracks?
 bibliography: cold-fronts.bib  
@@
 
-Cold fronts are a important feature of weather in extratropical regions and are partly responsible for the precipitation in these areas [@konstali2024].
+Cold fronts are an important feature of weather in extratropical regions and are partly responsible for the precipitation in these areas [@konstali2024].
 Cold fronts are also linked to extreme events, such as heavy precipitation [@catto2013] and severe wildfires in southern Australia [@cai2022].
@@
 To date, several computational methods have been developed that use different characteristics of fronts.
-Some of these methods looks for thermodynamic changes associated with temperature and humidity in different atmospheric levels [@berry2011], others identify changes in the wind field [@simmonds2012].
+Some of these methods look for thermodynamic changes associated with temperature and humidity in different atmospheric levels [@berry2011], others identify changes in the wind field [@simmonds2012].
 These methods rely on the data being in a regular grid (e.g.
@@
 
-For this study we use day-to-day changes in near-surface temperature (or drops in temperature) to estimate the passage of fronts [@reeder2015].
+For this study, we use day-to-day changes in near-surface temperature (or drops in temperature) to estimate the passage of fronts [@reeder2015].
 This very simple method can be applied to both data from station observations and gridded data (with varying resolution, up to 2.5 degrees [@king2024]).
@@
 As global temperatures continue to rise, understanding how this warming will impact the occurrence of cold fronts is very important.
-A few studies have addressed this question, for example @catto2014 used objective front identification [@berry2011] in 18 CMIP5 models and compared the changes between historical simulations between 1980-2005 and the Representative Concentration Pathway 8.5 between 2080-2100.
+A few studies have addressed this question, for example, @catto2014 used objective front identification [@berry2011] in 18 CMIP5 models and compared the changes between historical simulations between 1980-2005 and the Representative Concentration Pathway 8.5 between 2080-2100.
 They found an overall decrease in front frequency in the Northern Hemisphere, with a poleward shift in the maximum frequency and a significant decrease at high latitudes.
 In contrast, the Southern Hemisphere showed a poleward shift in the frequency maximum.
-On the other hand, @konstali2024, identified fronts in the CESM2-LE model simulations using gradient of equivalent potential temperature in 850 hPa and found an increase in the frequency of fronts that could be related to fronts becoming larger or more elongated in both hemispheres.
+On the other hand, @konstali2024, identified fronts in the CESM2-LE model simulations using the gradient of equivalent potential temperature in 850 hPa and found an increase in the frequency of fronts that could be related to fronts becoming larger or more elongated in both hemispheres.
 
-In this paper we build on the drops in temperature methodology to evaluate the change in frequency of cold fronts in a warmer future compared with the resent past using a set of models from CMIP6.
+In this paper we build on the drops in temperature methodology to evaluate the change in frequency of cold fronts in a warmer future compared with the recent past using a set of models from CMIP6.
 In an effort to understand possible causes to the potential changes, we explored the relationship between changes in front frequency and transient eddy kinetic energy, that is directly related to storm tracks.
@@
 
+For the analysis we use 10 models from phase 6 of the Coupled Model Intercomparison Project (CMIP6, @eyring2016), this includes Historical runs up to 2014 and the Shared Socioeconomic Pathways 5 (SSP5-8.5) from the Scenario Model Intercomparison Project (ScenarioMIP, @oneill2016) as described in @tbl-models.
+We choose these models based on the data available for the  historical and scenario experiments stored on the Australian National Computational Infrastructure (NCI) replica data store.
+
+While all models use the same nominal resolution of 100 km, we interpolate them to the same grid to easier the comparison.
+
 ```{r}
-#| label: model-table
+#| label: tbl-models
+#| tbl-cap: "List of the coupled model intercomparison project phase 6 (CMIP6) climate model used in this analysis."
+#| results: asis
 
@@
   unique(by = "source_id") |> 
+  _[order(source_id)] |> 
+  _[, let(references = c("@lovato2021; -@lovato2021a", "@consortiumec-earth2019; -@consortiumec-earth2019a",
+                         "@consortiumec-earth2020; -@consortiumec-earth2021", 
+                         "@consortiumec-earth2019b; -@consortiumec-earth2019c",
+                         "@consortiumec-earth2020a; -@consortiumec-earth2023",
+                         "@guo2018; -@guo2018a", "@volodin2019, @volodin2019a",
+                         "@volodin2019b; @volodin2019c", "@jungclaus2019; @schupfner2019",
+                         "@bentsen2019a; @bentsen2019b"))] |> 
+  _[, let(source_id = paste0(source_id, "\n[", references, "]"))] |> 
+  _[, .(source_id, institution_id, nominal_resolution)] |> 
   knitr::kable(col.names = c("Model", "Institution", "Original resolution"),
-               caption = "List of models used for the analysis.") |> 
-  kableExtra::kable_classic()
+               caption = "List of models used for the analysis.",
+               align = "llc",
+               format = "markdown", format.args = list(zero.print = "-"))
+
 ```
+
+### Front Identification
+
+We identify drops in temperature as a proxy of front passages by calculating the change in daily maximum temperature from one day to the next ($\Delta T$).
+The $\Delta T$ is calculated for all grid points and days over the studied periods.
+@king2024 defined a passage of a strong cold front as a $\Delta T \le -10°C$, meaning where the maximum temperature on the following day is at least 10$^{\circ}$C colder.
+While this value is reasonable for detecting strong drops in temperature over land, this can lead to an underestimation over sea due to the smaller temperature variability **(standard deviation map for max temp in supplemental material?)**.
+For this reason, in this study we use the value of the percentile 2.5th (as we want to capture the most negative temperature drops) as a threshold for each grid point to identify drops in temperature, ie $\Delta T \le 2.5^{th}$ percentile.
+The percentiles are calculated over the historical period between 1979 and 2014 for each model at each grid point.
+It is important to notice that by using this threshold we detect drops in temperature in the tropics that will not be associated with frontal systems.
+
+### Storm Tracks estimation
+
+To estimate the storm tracks we calculated the transient eddy kinetic energy (EKE) applying a Fast Fourier transform high-pass filter to the daily horizontal wind in 300 hPa with a 6-days cutoff for each model.
 
@@
 
+@fig-trend shows the change in $\Delta T$ frequency comparing the last 20 years of the 21st century (SSP5-8.5) and
+the last 20 years of the 20th century (historical run) in a) and the zonal mean for each latitude over sea b) and over land c).
+In the north hemisphere over land, there is a general decrease in temperature drops, particularly north to 40, with the exception of west and central Europe where there is an increase.
+This result is also observed by @catto2014 using CMIP5 and a different methodology to identify fronts and gives us confidence in the methodology used here.
+Over the oceans, there is also a decrease in the frequency of fronts that could be associated to a decrease in the meridional temperature gradient.
+The local maxima over the North Atlantic and negative values south to that area could indicate that the maximum frequency of $\Delta T$ is shifting toward the pole.
+
+In the South Hemisphere over land, the change is generally positive with almost 2% more events in some areas.
+Other authors have also found an increase in the frequency of fronts in the south and east regions of Africa and Australia [@catto2014, @konstali2024].
+@fig-trend a and b) show over the southern oceans a decrease in the frequency of $\Delta T$ between 30$^{\circ}$ and 45$^{\circ}$ S (depending on the longitude) and an increase to the south.
+This pattern indicates that the areas of higher front frequency will shift poleward in the SSP5-8.5.
+
 ```{r}
-#| label: trend
-#| fig-width: 10
-#| fig-height: 5
-#| fig.cap: "a) Multimodel change in drops in temperature frequency (%, SSP5-8.5 2018-2100 minus Historical 1980-2000) and zonal mean change over sea b) and land c), each grey line represent the models listed in Table 1 and in black, the multimodel mean.
-"
+#| label: fig-trend
+#| fig-width: 8
+#| fig-height: 4
+#| fig.cap: "a) Multimodel change in the frequency of $\\Delta T$ (days/year, SSP5-8.5 2018-2100 minus Historical 1980-2000) and zonal mean change over sea b) and land c), each grey line represents the models listed in Table 1 and in black, the multimodel mean.
+High topography areas are gray-shaded.
+Dotted areas show where at least 70% of the models agree in the sign of change"
 
@@
             var = meta[[1]][["var"]],
-            change = ssp585 - hist,
+            change = ssp585 / hist,
+            diff = ssp585 - hist,
             lat = round(lat, digits = 8),
@@
 
+topo <- rcdo::cdo_topo(grid = "r360x180") |> 
+  rcdo::cdo_execute(options = "-f nc") |> 
+  ReadNetCDF()
+
+sig <- data |> 
+  _[(model %in% models_in_gadi) & is.finite(change)] |>
+  dcast(lat + lon + model ~ var, value.var = "diff") |> 
+  _[, .(sig = sum(sign(deltap97.5), na.rm = TRUE)), by = .(lon, lat)]
+
 map_trend <- data |> 
   _[(model %in% models_in_gadi) & is.finite(change)] |>
-  dcast(lat + lon + model ~ var, value.var = "change") |> 
+  dcast(lat + lon + model ~ var, value.var = "diff") |> 
   _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lon, lat)] |> 
   ggplot(aes(ConvertLongitude(lon), lat)) +
-  geom_contour_fill(aes(z = mean_dt/365*100, fill = after_stat(level)),
-                    breaks = c(-Inf, seq(-3, 3, 0.2), Inf)) +
-  scale_fill_divergent_discretised(mid = "#f7f7f7",
+  geom_contour_fill(aes(z = mean_dt - 1, fill = after_stat(level)),
+                    breaks = c(-Inf, seq(-8, 8, 0.25), Inf)) +
+  scale_fill_divergent_discretised(name = "%", mid = "#f7f7f7",
                                    guide = guide_colorbar(barheight = 0.5),
@@
   geom_sf(data = global_map, inherit.aes = FALSE, fill = NA, linewidth = 0.2) +
+  geom_point(data = sig[is.cross(lon, lat, skip = 1) & abs(sig) >= 7], alpha = 0.3, shape = 4, size = 0.1) +
+  ggnewscale::new_scale_fill() +
+  geom_contour_fill(data = topo, aes(ConvertLongitude(lon), lat, z = topo), breaks = c(1300, Inf), fill = "grey90") +
   coord_sf(expand = FALSE) +
-  labs(x = NULL, y = NULL, fill = "%") +
+  labs(x = NULL, y = NULL, fill = "days/year") +
   theme_minimal(base_size = 12) +
@@
 
+annotation <- data.frame(ymin = -66, ymax = -56, xmin = -Inf, xmax = Inf, land = "Land")
+
 mean_trend <- data |> 
   _[(model %in% models_in_gadi) & is.finite(change)] |>
-  dcast(lat + lon + model ~ var, value.var = "change") |> 
+  dcast(lat + lon + model ~ var, value.var = "diff") |> 
   _[, let(land = MaskLand(lon, lat))] |> 
@@
   scale_y_latitude(breaks = seq(-80, 80 , 20)) +
+  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
+            fill = "#fbfbfb",
+            data = annotation,
+            inherit.aes = FALSE) +
   facet_wrap(~land) +
   tagger::tag_facets(tag_pool = c("b", "c")) +
-  labs(x = "Change (%)") +
+  labs(x = "days/year") +
   coord_fixed(ratio = 1/2.8) +
@@
 
+The mean time series of $\Delta T$ frequency in @fig-serie shows consistent trends.
+Interestingly, the trend around 50$^{\circ}$ N in the North Hemisphere is negative over sea and land since the beginning of the 20th century.
+In the South Hemisphere over the oceans, the time series for 30$^{\circ}$ and 40$^{\circ}$ S show a decrease of drops in temperature and the opposite happens on the 50$^{\circ}$ latitude band, indicating again a possible poleward shift of cold fronts in the future.
+It is important to notice the variability between models in the subfigure next to each panel.
+Each box plot is calculated using the year 2080-2100 for each model and region.
+The 50$^{\circ}$ latitude band in both hemispheres show the bigger changes but also the highest variability between models, however at least 70% of the models agree on the sign of the change.
+
 ```{r}
-#| label: serie
+#| label: fig-serie
 #| fig-width: 8
 #| fig-height: 5
-#| fig.cap: "Mean frequency of temperature drops for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different latitude bands.
+#| fig.cap: "Mean frequency of $\\Delta T$ for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different latitude bands.
 Left to each main figure the intermodel distribution is shown for the period 2080-2100."
@@
   _[!is.na(serie) & var == "deltap97.5" & model %in% models_in_gadi] |>
+  _[, .(mean_freq = weighted.mean(ssp585, cos(lat*pi/180))), by = .(serie, land, model)] |> 
   tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 
   setDT() |> 
-  _[, .(mean_freq = mean(ssp585)), by = .(lat, hemisphere, land, model)] |> 
   _[, let(scenario = "SSP5-8.5")]
@@
   _[!is.na(serie)] |> 
-  _[, .(mean_freq = mean(mean_freq)), by = .(land, serie, year, scenario)]
+  _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, year, scenario)]
 
@@
   setDT() |> 
-  # _[, let(lat = forcats::fct_relevel(lat, "55N", after = 0))] |> 
   ggplot(aes(year, mean_freq, color = lat, group = interaction(scenario, lat))) +
@@
   geom_ysideboxplot(data = model_spread, aes(x = lat), orientation = "x") +
-  scale_color_manual(values = c("#004C73", "#CC3311", "#EE7733")) +
+  scale_color_manual(values = c("#004C73", "#CC3311", "#ee8a33")) + 
   facet_grid(land~hemisphere, scale = "free") +
-  labs(x = "Year", y = "Frequency (days/year)", color = "Latitude") +
+  labs(x = "Year", y = latex2exp::TeX("$\\Delta T$ frequency (days/year)"), color = "Latitude") +
   theme_minimal() +
   theme(legend.position = "inside",
-        legend.position.inside = c(0.8, 0.9),
+        legend.position.inside = c(0.8, 0.93),
         legend.direction = "horizontal",
@@
 ```
+To explore the distribution of the change, @fig-distribution shows the change in frequency of $\Delta T$ using different percentiles as thresholds from 10 to 0.1th.
+The percentile values are also calculated from the historical period between 1979 and 2014 for each grid point and model as in @fig-trend.
+This figure shows that the change in the frequency of $\Delta T$ is bigger when the percentile is more extreme.
+For example in the North Hemisphere over sea, the 0.1 percentile shows that there will be less than half drops in temperature in the future over 50$^{\circ}$ N.
+However, it is important to notice that more extreme percentiles are associated to fewer events per year.
+According to this result, regions where we expect more drops in temperature in the future, e.g.
+land regions in the South Hemisphere, will also see more intense $\Delta T$, and the opposite for areas where the frequency will decreases.
 
 ```{r}
-#| label: distribution
+#| label: fig-distribution
 #| fig-width: 8
 #| fig-height: 5
-#| fig.cap: "Change in the frequency of drops in temperature (SSP5-8.5 2018-2100 minus Historical 1980-2000) for different percentiles calculates over the historical period (1979-2014)."
+#| fig.cap: "Fractional change in the frequency of $\\Delta T$ (SSP5-8.5 2018-2100 divided Historical 1980-2000) for different thresholds defined as the percentile value for each grid point and model calculated over the historical period (1979-2014)."
 
@@
   _[!is.na(serie)] |> 
-  _[, .(mean_freq = mean(mean_freq)), by = .(land, serie, percentile, scenario)] |> 
+  _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, percentile, scenario)] |> 
   dcast(land + serie + percentile ~ scenario, value.var = "mean_freq") |> 
@@
   setDT() |> 
-  ggplot(aes(percentile, ssp585 / historical)) +
+  ggplot(aes(factor(round(100-as.numeric(percentile), digits = 1)), ssp585 / historical)) +
   geom_col(aes(fill = lat), position = "dodge") +
-  scale_fill_manual(values = c("#004C73", "#CC3311", "#EE7733")) +
+  scale_fill_manual(values = c("#004C73", "#CC3311", "#ee8a33")) +
   scale_y_log10() +
   facet_grid(land~hemisphere) +
-  labs(x = "Percentile", y = "Ratio (SSP5-8.5/Historical)",
+  labs(x = "Percentile", y = latex2exp::TeX("Fractional change of $\\Delta T$"),
        fill = "Latitude") +
@@
   theme(legend.position = "inside",
-        legend.position.inside = c(0.8, 0.9),
+        legend.position.inside = c(0.8, 0.93),
         legend.direction = "horizontal",
@@
 
+@fig-eke explores the relationship between the change in EKE and frequency of $\Delta T$ comparing the scenario SSP5-8.5 (2080-2100) and the recent past (1980-2000).
+In the North Hemisphere, both changes in EKE and in the frequency of $\Delta T$ are mainly negative (@fig-eke a).
+The zonal mean shows that this positive relationship holds well over oceans with negative changes of $\Delta T$ frequency and EKE around 40$^{\circ}$ N (@fig-eke b).
+At first glance, the relationship in the South Hemisphere is negative due mainly to the mean changes over land where the $\Delta T$ increases while the EKE shifts to the south.
+**why?
+are we detecting changes in temp not related to cyclones and frontal things?** Over the sea, the changes are consistent with a poleward shift of the storm tracks (already reported by @ogorman2010, @tamarin2017, @chemke2022) and the maximum frequency of $\Delta T$.
+
 ```{r}
-#| label: eke
+#| label: fig-eke
 #| fig-width: 8
 #| fig-height: 4
-#| fig.cap: "Changes in EKE and drops in temperature frequency for SSP5-8.5 compared with the historical period and each model.
-Values are calculated for northern () and southern () hemisphere and over sea (circles) and land(triangles).
-Each solid line correspond to a linea regresion for each hemisphere.
-"
+#| fig.cap: "Mean change in EKE and $\\Delta T$ frequency for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model.
+Values are calculated for the northern (25$^{\\circ}$ to 55$^{\\circ}$) and southern (-25$^{\\circ}$ to -55$^{\\circ}$) hemispheres and over sea (circles) and land (triangles).
+Each solid line corresponds to a linear regression for each hemisphere.
+b) and c) show the zonal mean for EKE ($m^2 s^2$) and $\\Delta T$ (days/year) over land and sea respectively"
 
@@
   _[(model %in% models_in_gadi) & is.finite(change)] |>
-  dcast(lat + lon + model ~ var, value.var = "change") |>
+  dcast(lat + lon + model ~ var, value.var = "diff") |>
 
@@
         mean_deltat = mean(deltap97.5)), by = .(region, land, model)] |> 
-  ggplot(aes(mean_eke, mean_deltat)) +
+  ggplot(aes((mean_eke - 1)*100, (mean_deltat - 1)*100)) +
   geom_point(aes(color = region, shape = land), size = 2) +
@@
   scale_color_manual(values = c("cyan4", "orange")) +
-  labs(x = "EKE change", y = "Temp drop frequency change",
+  labs(x = latex2exp::TeX("EKE change ($m^2 s^2$)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
        color = NULL, shape = NULL) +
@@
   theme(legend.position = "inside",
-        legend.position.inside = c(0.85, 0.15),
+        legend.position.inside = c(0.85, 0.1),
         legend.direction = "horizontal",
@@
   _[(model %in% models_in_gadi) & is.finite(change)] |>
-  dcast(lat + lon + model ~ var, value.var = "change") |> 
+  dcast(lat + lon + model ~ var, value.var = "diff") |> 
   _[, let(land = MaskLand(lon, lat))] |> 
   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land)]
+  _[, .(mean_dt = mean(deltap97.5-1, na.rm = TRUE)), by = .(lat, land)]
 
@@
   _[(model %in% models_in_gadi) & is.finite(change)] |>
-  dcast(lat + lon + model ~ var, value.var = "change") |> 
+  dcast(lat + lon + model ~ var, value.var = "diff") |> 
   _[, let(land = MaskLand(lon, lat))] |> 
@@
   _[, .(mean_eke = mean(eke, na.rm = TRUE)), by = .(lat, land, model)] |> 
-  ggplot(aes(mean_eke, lat)) +
+  ggplot(aes(mean_eke-1, lat)) +
   geom_hline(yintercept = 0, color = "grey80") +
@@
   geom_path(data = deltat_mean, aes(x = mean_dt, color = "DeltaT")) +
-  scale_color_manual(values = c("DeltaT" = "red4", "EKE" = "black")) +
+  scale_color_manual(values = c("DeltaT" = "red4", "EKE" = "black"),
+                     labels = c(latex2exp::TeX("$\\Delta T$"), "EKE")) +
+  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
+            fill = "#fbfbfb",
+            data = annotation,
+            inherit.aes = FALSE) +
   scale_y_latitude(breaks = seq(-80, 80 , 20)) +
@@
   tagger::tag_facets(tag_pool = c("b", "c")) +
-  labs(x = "Change (%)", color = NULL) +
-  coord_fixed(ratio = 1/1.5) +
+  labs(x = "Change", color = NULL) +
   theme_minimal() +
@@
         strip.text.x = element_text(size = 12),
-        plot.margin = unit(c(0, 0 , 1.9, -0.40), units = "cm"),
+        plot.margin = unit(c(0, 0 , 0, -0.40), units = "cm"),
         tagger.panel.tag.text = element_text(size = 12),
@@
 
-Detection of fronts over oceans when using a constant threshold
+What we did
+
+Summary of results
+
+Detection of fronts over oceans when using a constant threshold underestimates the fronts, using a percentile means detecting a similar number of fronts in each grid point.
 Weak fronts are not detected? 
+Contrast land - sea play a role here?
+Why EKE does not correlate well with deltaT in some areas? particularly SH-Land?
 
@@
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
+CLEX, NCI, etc
 
@@
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
+The data used in the manuscript is publicly available for CMIP6 data at the Earth System Grid  Federation [https://esgf-node.llnl.gov/projects/cmip6/](https://esgf-node.llnl.gov/projects/cmip6/).
+A version-controlled repository of the code used to create this analysis, including the necessary code to download the derived data can be found at <https://github.com/paocorrales/t-drop-trends>.
+The derived data that support the findings of this study are also openly available in Zenodo at http://doi.org/, version 0.xx.
 

### 2025-02-10 article/article.qmd
fix more text

@@
 These methods rely on the data being in a regular grid (e.g.
-outputs from numerical models or reanalysis) and it results depend on setting different thresholds that need to be calibrated for different model resolutions.
+outputs from numerical models or reanalysis) and their results depend on setting different thresholds, which need to be calibrated for different model resolutions.
 
@@
 This very simple method can be applied to both data from station observations and gridded data (with varying resolution, up to 2.5 degrees [@king2024]).
-Previous studies have shown that this method is suitable for the detection of strong cold fronts in southeastern Australia [@cai2022, @reeder2015].
+Previous studies have shown that this method is suitable for the detection of strong cold fronts in southeastern Australia [@cai2022; @reeder2015].
 Recently, @king2024 demonstrated that this method can successfully identify cold fronts with results comparable to those obtained with other methodologies.
@@
 In an effort to understand possible causes to the potential changes, we explored the relationship between changes in front frequency and transient eddy kinetic energy, that is directly related to storm tracks.
-Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@ogorman2010, @tamarin2017, @chemke2022], and these changes could potentially influence the location and intensity of future cold fronts.
+Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@ogorman2010; @tamarin2017; @chemke2022], and these changes could potentially influence the location and intensity of future cold fronts.
 
@@
 For the analysis we use 10 models from phase 6 of the Coupled Model Intercomparison Project (CMIP6, @eyring2016), this includes Historical runs up to 2014 and the Shared Socioeconomic Pathways 5 (SSP5-8.5) from the Scenario Model Intercomparison Project (ScenarioMIP, @oneill2016) as described in @tbl-models.
-We choose these models based on the data available for the  historical and scenario experiments stored on the Australian National Computational Infrastructure (NCI) replica data store.
+We selected these models based on the data available for the  historical and scenario experiments stored on the Australian National Computational Infrastructure (NCI) replica data store.
 
-While all models use the same nominal resolution of 100 km, we interpolate them to the same grid to easier the comparison.
+While all models have a nominal resolution of 100 km, we bilinear interpolated them onto the same grid to facilitate comparison.
 
@@
 The $\Delta T$ is calculated for all grid points and days over the studied periods.
-@king2024 defined a passage of a strong cold front as a $\Delta T \le -10°C$, meaning where the maximum temperature on the following day is at least 10$^{\circ}$C colder.
-While this value is reasonable for detecting strong drops in temperature over land, this can lead to an underestimation over sea due to the smaller temperature variability **(standard deviation map for max temp in supplemental material?)**.
+@king2024 defined a passage of a strong cold front as a $\Delta T \le -10^{\circ}C$, meaning where the maximum temperature on the following day is at least 10$^{\circ}$C colder.
+While this value is reasonable for detecting strong drops in temperature over land, it may underestimate drops over the ocean due to the smaller temperature variability **(standard deviation map for max temp in supplemental material?)**.
 For this reason, in this study we use the value of the percentile 2.5th (as we want to capture the most negative temperature drops) as a threshold for each grid point to identify drops in temperature, ie $\Delta T \le 2.5^{th}$ percentile.
+We chose this percentile threshold as the $\Delta T$ mean of values associated over land for midlatitudes (25$^{\circ}$ to 55$^{\circ}$) is close to $-10^{\circ}C$, the threshold used by @king2024.
 The percentiles are calculated over the historical period between 1979 and 2014 for each model at each grid point.
-It is important to notice that by using this threshold we detect drops in temperature in the tropics that will not be associated with frontal systems.
+It is important to note that by using this threshold we detect drops in temperature in the tropics that are unrelated to frontal systems.
+For this reason, the analysis will focus on mid and high latitudes.
 
@@
 
-To estimate the storm tracks we calculated the transient eddy kinetic energy (EKE) applying a Fast Fourier transform high-pass filter to the daily horizontal wind in 300 hPa with a 6-days cutoff for each model.
+We estimate storm tracks by calculating transient eddy kinetic energy (EKE) using a Fast Fourier Transform high-pass filter applied to daily horizontal wind at 300 hPa, with a 6-day cutoff for each model.
 
@@
 the last 20 years of the 20th century (historical run) in a) and the zonal mean for each latitude over sea b) and over land c).
-In the north hemisphere over land, there is a general decrease in temperature drops, particularly north to 40, with the exception of west and central Europe where there is an increase.
-This result is also observed by @catto2014 using CMIP5 and a different methodology to identify fronts and gives us confidence in the methodology used here.
+In the Northern Hemisphere over land, there is a general decrease in temperature drops, particularly north of 40$^{\circ}$ latitude, except in western and central Europe where there is an increase.
+This result is also observed by @catto2014 using CMIP5 and a different methodology to identify fronts reinforcing the reliability of the methodology used here.
 Over the oceans, there is also a decrease in the frequency of fronts that could be associated to a decrease in the meridional temperature gradient.
@@
 
-In the South Hemisphere over land, the change is generally positive with almost 2% more events in some areas.
-Other authors have also found an increase in the frequency of fronts in the south and east regions of Africa and Australia [@catto2014, @konstali2024].
+In the Southern Hemisphere over land, the change is generally positive with 5 events per year in some areas.
+Other authors have also found an increase in the frequency of fronts in the south and east regions of Africa and Australia [@catto2014; @konstali2024].
 @fig-trend a and b) show over the southern oceans a decrease in the frequency of $\Delta T$ between 30$^{\circ}$ and 45$^{\circ}$ S (depending on the longitude) and an increase to the south.
 This pattern indicates that the areas of higher front frequency will shift poleward in the SSP5-8.5.
+
+**Increase over land vs decrease over sea could be related also to the thermal inertia / land get hotter?**
 
@@
   ggplot(aes(ConvertLongitude(lon), lat)) +
-  geom_contour_fill(aes(z = mean_dt - 1, fill = after_stat(level)),
+  geom_contour_fill(aes(z = mean_dt, fill = after_stat(level)),
                     breaks = c(-Inf, seq(-8, 8, 0.25), Inf)) +
-  scale_fill_divergent_discretised(name = "%", mid = "#f7f7f7",
+  scale_fill_divergent_discretised(name = "days/year", mid = "#f7f7f7",
                                    guide = guide_colorbar(barheight = 0.5),
@@
 
-The mean time series of $\Delta T$ frequency in @fig-serie shows consistent trends.
-Interestingly, the trend around 50$^{\circ}$ N in the North Hemisphere is negative over sea and land since the beginning of the 20th century.
-In the South Hemisphere over the oceans, the time series for 30$^{\circ}$ and 40$^{\circ}$ S show a decrease of drops in temperature and the opposite happens on the 50$^{\circ}$ latitude band, indicating again a possible poleward shift of cold fronts in the future.
-It is important to notice the variability between models in the subfigure next to each panel.
+The mean time series of $\Delta T$ frequency in @fig-serie shows trends consistent with @fig-trend a).
+Interestingly, the trend around 50$^{\circ}$ N in the Northern Hemisphere is negative over sea and land since the beginning of the 20th century.
+In the Southern Hemisphere over the oceans, the time series for 30$^{\circ}$ and 40$^{\circ}$ S show a decrease of drops in temperature and the opposite happens on the 50$^{\circ}$ latitude band, indicating again a possible poleward shift of cold fronts in the future.
+It is important to note the variability between models in the subfigure next to each panel.
 Each box plot is calculated using the year 2080-2100 for each model and region.
@@
 The percentile values are also calculated from the historical period between 1979 and 2014 for each grid point and model as in @fig-trend.
-This figure shows that the change in the frequency of $\Delta T$ is bigger when the percentile is more extreme.
-For example in the North Hemisphere over sea, the 0.1 percentile shows that there will be less than half drops in temperature in the future over 50$^{\circ}$ N.
-However, it is important to notice that more extreme percentiles are associated to fewer events per year.
-According to this result, regions where we expect more drops in temperature in the future, e.g.
-land regions in the South Hemisphere, will also see more intense $\Delta T$, and the opposite for areas where the frequency will decreases.
+This figure shows that the change in the frequency of $\Delta T$ is bigger at lower (more extreme) percentile thresholds.
+For example in the Northern Hemisphere over sea, the 0.1 percentile threshold shows that there will be less than half drops in temperature in the future over 50$^{\circ}$ N.
+However, it is important to note that lower percentiles thresholds are associated to rarer events, meaning fewer occurrences per year.
+These results suggest that regions projected to experience more temperature drops in the future, such as land areas in midlatitudes in the Southern Hemisphere, will also experience more intense $\Delta T$, whereas regions with a decreasing frequency of events will see less intense drops.
+
+**why? **
 
@@
 @fig-eke explores the relationship between the change in EKE and frequency of $\Delta T$ comparing the scenario SSP5-8.5 (2080-2100) and the recent past (1980-2000).
-In the North Hemisphere, both changes in EKE and in the frequency of $\Delta T$ are mainly negative (@fig-eke a).
+In the Northern Hemisphere, both EKE and the frequency of $\Delta T$ show mainly negative changes (@fig-eke a).
 The zonal mean shows that this positive relationship holds well over oceans with negative changes of $\Delta T$ frequency and EKE around 40$^{\circ}$ N (@fig-eke b).
-At first glance, the relationship in the South Hemisphere is negative due mainly to the mean changes over land where the $\Delta T$ increases while the EKE shifts to the south.
+In the Southern Hemisphere, the overall relationship appears negative, primarily due to increasing $\Delta T$ frequency over land while EKE maxima shifts poleward.
 **why?

### 2025-02-12 article/article.qmd
Change small details in figures

@@
 High topography areas are gray-shaded.
-Dotted areas show where at least 70% of the models agree in the sign of change"
+Dotted areas show where at least 90% of the models agree in the sign of change"
 
@@
   geom_contour_fill(aes(z = mean_dt, fill = after_stat(level)),
-                    breaks = c(-Inf, seq(-8, 8, 0.25), Inf)) +
+                    breaks = c(-Inf, seq(-8, 8, 0.5), Inf)) +
   scale_fill_divergent_discretised(name = "days/year", mid = "#f7f7f7",
@@
   geom_sf(data = global_map, inherit.aes = FALSE, fill = NA, linewidth = 0.2) +
-  geom_point(data = sig[is.cross(lon, lat, skip = 1) & abs(sig) >= 7], alpha = 0.3, shape = 4, size = 0.1) +
+  geom_point(data = sig[is.cross(lon, lat, skip = 1) & abs(sig) >= 9], alpha = 0.3, shape = 4, size = 0.1) +
   ggnewscale::new_scale_fill() +
@@
         axis.text.y = element_blank(),
-        axis.ticks.y = element_blank(), plot.margin = unit(c(0, 0 , 0 , 0), units = "mm"))
+        axis.ticks.y = element_blank(), 
+        plot.margin = unit(c(0, -0.4 , 0 , 0), units = "cm"))
 
@@
   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50 - NH",
-                        lat %between% c(38, 40) & land == "Sea", "40 - NH",
-                        lat %between% c(28, 30) & land == "Sea", "30 - NH", 
-                        lat %between% c(48, 50) & land == "Land", "50 - NH",
-                        lat %between% c(38, 40) & land == "Land", "40 - NH",
-                        lat %between% c(28, 30) & land == "Land", "30 - NH",
-                        lat %between% c(-50, -48) & land == "Land", "50 - SH",
-                        lat %between% c(-40, -38) & land == "Land", "40 - SH",
-                        lat %between% c(-30, -28) & land == "Land", "30 - SH",
-                        lat %between% c(-50, -48) & land == "Sea", "50 - SH",
-                        lat %between% c(-40, -38) & land == "Sea", "40 - SH",
-                        lat %between% c(-30, -28) & land == "Sea", "30 - SH"))] |>
+  _[, let(serie = fcase(lat %between% c(48, 50), "50 - NH",
+                        lat %between% c(38, 40), "40 - NH",
+                        lat %between% c(28, 30), "30 - NH", 
+                        lat %between% c(-50, -48), "50 - SH",
+                        lat %between% c(-40, -38), "40 - SH",
+                        lat %between% c(-30, -28), "30 - SH"))] |>
   _[!is.na(serie) & var == "deltap97.5" & model %in% models_in_gadi] |>
@@
   setDT() |> 
-  _[, let(scenario = "SSP5-8.5")]
+  _[, let(scenario = "SSP5-8.5")] |> 
+  _[!(land == "Land" & hemisphere == "SH" & lat == 50)]
 
@@
   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50 - NH",
-                        lat %between% c(38, 40) & land == "Sea", "40 - NH",
-                        lat %between% c(28, 30) & land == "Sea", "30 - NH", 
-                        lat %between% c(48, 50) & land == "Land", "50 - NH",
-                        lat %between% c(38, 40) & land == "Land", "40 - NH",
-                        lat %between% c(28, 30) & land == "Land", "30 - NH",
-                        lat %between% c(-50, -48) & land == "Land", "50 - SH",
-                        lat %between% c(-40, -38) & land == "Land", "40 - SH",
-                        lat %between% c(-30, -28) & land == "Land", "30 - SH",
-                        lat %between% c(-50, -48) & land == "Sea", "50 - SH",
-                        lat %between% c(-40, -38) & land == "Sea", "40 - SH",
-                        lat %between% c(-30, -28) & land == "Sea", "30 - SH"))] |>
+  _[, let(serie = fcase(lat %between% c(48, 50), "50 - NH",
+                        lat %between% c(38, 40), "40 - NH",
+                        lat %between% c(28, 30), "30 - NH", 
+                        lat %between% c(-50, -48), "50 - SH",
+                        lat %between% c(-40, -38), "40 - SH",
+                        lat %between% c(-30, -28), "30 - SH"))] |>
   _[!is.na(serie)] |> 
-  _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, year, scenario)]
+  _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, year, scenario)] |> 
+  _[!(land == "Land" & serie == "50 - SH")]
 
@@
   setDT() |> 
+  _[year >= 1900] |> 
   ggplot(aes(year, mean_freq, color = lat, group = interaction(scenario, lat))) +
@@
   theme(legend.position = "inside",
-        legend.position.inside = c(0.8, 0.93),
+        legend.position.inside = c(0.78, 0.93),
         legend.direction = "horizontal",
@@
         strip.text = element_text(size = 12),
-        ggside.axis.text.x = element_blank())
+        ggside.axis.text.x = element_blank(),
+        ggside.panel.scale = 0.2)
 
@@
 #| fig.cap: "Mean change in EKE and $\\Delta T$ frequency for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model.
-Values are calculated for the northern (25$^{\\circ}$ to 55$^{\\circ}$) and southern (-25$^{\\circ}$ to -55$^{\\circ}$) hemispheres and over sea (circles) and land (triangles).
+Values are calculated for the northern (25$^{\\circ}$ to 65$^{\\circ}$) and southern (-25$^{\\circ}$ to -65$^{\\circ}$) hemispheres and over sea (circles) and land (triangles).
 Each solid line corresponds to a linear regression for each hemisphere.
-b) and c) show the zonal mean for EKE ($m^2 s^2$) and $\\Delta T$ (days/year) over land and sea respectively"
+b) and c) show the zonal mean for EKE ($m^2 s^{-2}$) and $\\Delta T$ (days/year) over land and sea respectively"
 
@@
   _[, let(land = MaskLand(lon, lat),
-          region = fcase(lat %between% c(25, 55), "NH",
-                         lat %between% c(-55, -25), "SH",
+          region = fcase(lat %between% c(25, 65), "NH",
+                         lat %between% c(-65, -25), "SH",
                          default = NA))] |> 
@@
         mean_deltat = mean(deltap97.5)), by = .(region, land, model)] |> 
-  ggplot(aes((mean_eke - 1)*100, (mean_deltat - 1)*100)) +
+  ggplot(aes(mean_eke, mean_deltat)) +
   geom_point(aes(color = region, shape = land), size = 2) +
@@
   scale_color_manual(values = c("cyan4", "orange")) +
-  labs(x = latex2exp::TeX("EKE change ($m^2 s^2$)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
+  labs(x = latex2exp::TeX("EKE change ($m^2 s^{-2}$)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
        color = NULL, shape = NULL) +
@@
   theme(legend.position = "inside",
-        legend.position.inside = c(0.85, 0.1),
+        legend.position.inside = c(0.83, 0.08),
         legend.direction = "horizontal",
         legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
-        legend.margin = margin(0, 0, 0, 0, unit = "mm"),
+        legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
         panel.background = element_rect(fill = "#fbfbfb", color = NA))
@@
             inherit.aes = FALSE) +
+  annotate(geom = "rect", ymin = c(-90, -20, 90), ymax = c(-65, 20, 65), xmin = -Inf, xmax = Inf,
+           fill = "grey93", alpha = 0.7) +
   scale_y_latitude(breaks = seq(-80, 80 , 20)) +
@@
         plot.margin = unit(c(0, 0 , 0, -0.40), units = "cm"),
-        tagger.panel.tag.text = element_text(size = 12),
-        tagger.panel.tag.background = element_rect(fill = "white", color = "white"))
+        tagger.panel.tag.text = element_text(size = 12))
 

### 2025-02-17 article/article.qmd
weighted mean

@@
 
-We estimate storm tracks by calculating transient eddy kinetic energy (EKE) using a Fast Fourier Transform high-pass filter applied to daily horizontal wind at 300 hPa, with a 6-day cutoff for each model.
+To estimate the storm track location we use the transient eddy kinetic energy (EKE), used by many others to estimate storm tracks [@ogorman2010; @chemke2022].
+We calculated EKE using a Fast Fourier Transform high-pass filter applied to daily horizontal wind at 300 hPa, with a 6-day cutoff for each model.
 
@@
 #| fig-height: 5
-#| fig.cap: "Mean frequency of $\\Delta T$ for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different latitude bands.
+#| fig.cap: "Multimodel mean frequency of $\\Delta T$ for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different latitude bands.
 Left to each main figure the intermodel distribution is shown for the period 2080-2100."
@@
   na.omit() |> 
-  _[, .(mean_eke = mean(eke),
-        mean_deltat = mean(deltap97.5)), by = .(region, land, model)] |> 
+  _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180)),
+        mean_deltat = weighted.mean(deltap97.5, w = cos(lat*pi/180))), by = .(region, land, model)] |> 
   ggplot(aes(mean_eke, mean_deltat)) +
@@
   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, .(mean_dt = mean(deltap97.5-1, na.rm = TRUE)), by = .(lat, land)]
+  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land)]
 
@@
   _[, .(mean_eke = mean(eke, na.rm = TRUE)), by = .(lat, land, model)] |> 
-  ggplot(aes(mean_eke-1, lat)) +
+  ggplot(aes(mean_eke, lat)) +
   geom_hline(yintercept = 0, color = "grey80") +

### 2025-02-18 article/article.qmd
add meridional gradient of temperature change

@@
                     "EC-Earth3-Veg-LR", "GFDL-CM4")
+
+data <- purrr::map(Sys.glob(here("data/eke_mean/*hist*mean.nc")), function(f) {
+
+  message(f)
+  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_{scenario}_mean.nc", "{var}_{model}_{scenario}_mean.nc"))
+
+  if (meta[[1]][["var"]] == "eke") {
+    var <- "ua"
+  } else {
+    var <- "tasmax"
+  }
+  file_scenario <- here(paste0("data/eke_mean/", meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585_mean.nc"))
+
+  ReadNetCDF(f, vars = c(hist = var)) |> 
+    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
+    _[, let(model = meta[[1]][["model"]],
+            var = meta[[1]][["var"]],
+            change = ssp585 / hist,
+            diff = ssp585 - hist,
+            lat = round(lat, digits = 8),
+            plev = NULL)] |> 
+    _[]
+
+}) |> rbindlist()
+
+tas <- purrr::map(Sys.glob(here("data/tas/*hist*.nc")), function(f) {
+
+  message(f)
+  meta <- unglue::unglue(basename(f), patterns = c("{var}_day_{model}_{scenario}_{member}_{grid}_20150101-21001231.nc"))
+
+  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
+                                        meta[[1]][["var"]], "_day_", meta[[1]][["model"]], "_ssp585*")))
+
+  var <- meta[[1]][["var"]]
+
+  ReadNetCDF(f, vars = c(hist = var)) |> 
+    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
+    _[, let(model = meta[[1]][["model"]],
+            var = "dt_dy",
+            lat = round(lat, digits = 8),
+            hist = metR:::.derv(hist, lat)*(180/pi/6371000),
+            ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
+    _[, let(change = ssp585 / hist,
+            diff = ssp585 - hist)]
+
+}) |> rbindlist()
+
+data <- rbind(data, tas) |> 
+  _[, let(land = MaskLand(lon, lat))] |> 
+  _[, let(land = factor(land, labels = c("Sea", "Land")))]
+
 ```
@@
 
-data <- purrr::map(Sys.glob(here("data/eke_mean/*hist*mean.nc")), function(f) {
-
-  message(f)
-  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_{scenario}_mean.nc", "{var}_{model}_{scenario}_mean.nc"))
-
-  if (meta[[1]][["var"]] == "eke") {
-    var <- "ua"
-  } else {
-    var <- "tasmax"
-  }
-  file_scenario <- here(paste0("data/eke_mean/", meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585_mean.nc"))
-
-  ReadNetCDF(f, vars = c(hist = var)) |> 
-    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
-    _[, let(model = meta[[1]][["model"]],
-            var = meta[[1]][["var"]],
-            change = ssp585 / hist,
-            diff = ssp585 - hist,
-            lat = round(lat, digits = 8),
-            plev = NULL)] |> 
-    _[]
-
-}) |> rbindlist()
-
 topo <- rcdo::cdo_topo(grid = "r360x180") |> 
@@
 sig <- data |> 
-  _[(model %in% models_in_gadi) & is.finite(change)] |>
+  _[(model %in% models_in_gadi) & var == "deltap97.5"] |>
   dcast(lat + lon + model ~ var, value.var = "diff") |> 
@@
 mean_trend <- data |> 
-  _[(model %in% models_in_gadi) & is.finite(change)] |>
+  _[(model %in% models_in_gadi) & var == "deltap97.5"] |>
   dcast(lat + lon + model ~ var, value.var = "diff") |> 
@@
 model_spread <- data |> 
-  _[, let(land = MaskLand(lon, lat))] |> 
-  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
   _[, let(serie = fcase(lat %between% c(48, 50), "50 - NH",
@@
 #| fig-height: 4
-#| fig.cap: "Mean change in EKE and $\\Delta T$ frequency for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model.
+#| fig.cap: "Mean change in $\\Delta T$ frequency and a) $\\nabla _{y}$ and b) EKE for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model.
 Values are calculated for the northern (25$^{\\circ}$ to 65$^{\\circ}$) and southern (-25$^{\\circ}$ to -65$^{\\circ}$) hemispheres and over sea (circles) and land (triangles).
-Each solid line corresponds to a linear regression for each hemisphere.
-b) and c) show the zonal mean for EKE ($m^2 s^{-2}$) and $\\Delta T$ (days/year) over land and sea respectively"
+Each solid line corresponds to a linear regression for each hemisphere."
 
 change <- data |> 
-  _[(model %in% models_in_gadi) & is.finite(change)] |>
-  dcast(lat + lon + model ~ var, value.var = "diff") |>
+  _[(model %in% models_in_gadi)] |>
+  dcast(lat + lon + model + land ~ var, value.var = "diff") |>
 
-  _[, let(land = MaskLand(lon, lat),
-          region = fcase(lat %between% c(25, 65), "NH",
+  _[, let(region = fcase(lat %between% c(25, 65), "NH",
                          lat %between% c(-65, -25), "SH",
                          default = NA))] |> 
-  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  na.omit() |> 
-  _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180)),
-        mean_deltat = weighted.mean(deltap97.5, w = cos(lat*pi/180))), by = .(region, land, model)] |> 
+  _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180), na.rm = TRUE),
+        mean_dt_dy = weighted.mean(dt_dy, w = cos(lat*pi/180), na.rm = TRUE),
+        mean_deltat = weighted.mean(deltap97.5, w = cos(lat*pi/180), na.rm = TRUE)), by = .(region, land, model)] |> 
+  na.omit()
+
+gradt <- change |> 
+  _[, let(mean_dt_dy = fifelse(region == "NH", -mean_dt_dy, mean_dt_dy))] |> 
+  ggplot(aes(mean_dt_dy*1000, mean_deltat)) +
+  geom_point(aes(color = region, shape = land), size = 2) +
+  geom_smooth(aes(colour = region), method = "lm", se = FALSE) +
+  scale_color_manual(values = c("cyan4", "orange")) +
+  labs(x = latex2exp::TeX("$\\nabla _{y} T$ change (K/km)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
+       color = NULL, shape = NULL) +
+  theme_minimal() +
+  theme(legend.position = "inside",
+        legend.position.inside = c(0.83, 0.08),
+        legend.direction = "horizontal",
+        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
+        legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
+        panel.background = element_rect(fill = "#fbfbfb", color = NA))
+
+eke <- change |> 
   ggplot(aes(mean_eke, mean_deltat)) +
@@
 
-deltat_mean <- data |> 
-  _[(model %in% models_in_gadi) & is.finite(change)] |>
-  dcast(lat + lon + model ~ var, value.var = "diff") |> 
-  _[, let(land = MaskLand(lon, lat))] |> 
-  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land)]
+# deltat_mean <- data |> 
+#   _[(model %in% models_in_gadi) & is.finite(change)] |>
+#   dcast(lat + lon + model ~ var, value.var = "diff") |> 
+#   _[, let(land = MaskLand(lon, lat))] |> 
+#   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
+#   _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land)]
+# 
+# mean_trend_dt <- data |> 
+#   _[(model %in% models_in_gadi) & is.finite(change)] |>
+#   dcast(lat + lon + model ~ var, value.var = "diff") |> 
+#   _[, let(land = MaskLand(lon, lat))] |> 
+#   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
+#   _[, .(mean_eke = mean(eke, na.rm = TRUE)), by = .(lat, land, model)] |> 
+#   ggplot(aes(mean_eke, lat)) +
+#   geom_hline(yintercept = 0, color = "grey80") +
+#   geom_vline(xintercept = 0, color = "grey80") +
+#   geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
+#   stat_summary(orientation = "y", geom = "line", aes(color = "EKE")) +
+#   geom_path(data = deltat_mean, aes(x = mean_dt, color = "DeltaT")) +
+#   scale_color_manual(values = c("DeltaT" = "red4", "EKE" = "black"),
+#                      labels = c(latex2exp::TeX("$\\Delta T$"), "EKE")) +
+#   geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
+#             fill = "#fbfbfb",
+#             data = annotation,
+#             inherit.aes = FALSE) +
+#   annotate(geom = "rect", ymin = c(-90, -20, 90), ymax = c(-65, 20, 65), xmin = -Inf, xmax = Inf,
+#            fill = "grey93", alpha = 0.7) +
+#   scale_y_latitude(breaks = seq(-80, 80 , 20)) +
+#   facet_wrap(~land) +
+#   tagger::tag_facets(tag_pool = c("b", "c")) +
+#   labs(x = "Change", color = NULL) +
+#   theme_minimal() +
+#   theme(legend.position = "bottom",
+#         panel.background = element_rect(fill = "#fbfbfb", color = NA),
+#         strip.text.x = element_text(size = 12),
+#         plot.margin = unit(c(0, 0 , 0, -0.40), units = "cm"),
+#         tagger.panel.tag.text = element_text(size = 12))
 
-mean_trend_dt <- data |> 
-  _[(model %in% models_in_gadi) & is.finite(change)] |>
-  dcast(lat + lon + model ~ var, value.var = "diff") |> 
-  _[, let(land = MaskLand(lon, lat))] |> 
-  _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, .(mean_eke = mean(eke, na.rm = TRUE)), by = .(lat, land, model)] |> 
-  ggplot(aes(mean_eke, lat)) +
-  geom_hline(yintercept = 0, color = "grey80") +
-  geom_vline(xintercept = 0, color = "grey80") +
-  geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
-  stat_summary(orientation = "y", geom = "line", aes(color = "EKE")) +
-  geom_path(data = deltat_mean, aes(x = mean_dt, color = "DeltaT")) +
-  scale_color_manual(values = c("DeltaT" = "red4", "EKE" = "black"),
-                     labels = c(latex2exp::TeX("$\\Delta T$"), "EKE")) +
-  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
-            fill = "#fbfbfb",
-            data = annotation,
-            inherit.aes = FALSE) +
-  annotate(geom = "rect", ymin = c(-90, -20, 90), ymax = c(-65, 20, 65), xmin = -Inf, xmax = Inf,
-           fill = "grey93", alpha = 0.7) +
-  scale_y_latitude(breaks = seq(-80, 80 , 20)) +
-  facet_wrap(~land) +
-  tagger::tag_facets(tag_pool = c("b", "c")) +
-  labs(x = "Change", color = NULL) +
-  theme_minimal() +
-  theme(legend.position = "bottom",
-        panel.background = element_rect(fill = "#fbfbfb", color = NA),
-        strip.text.x = element_text(size = 12),
-        plot.margin = unit(c(0, 0 , 0, -0.40), units = "cm"),
-        tagger.panel.tag.text = element_text(size = 12))
-
-plot_grid(change, mean_trend_dt, labels = c('a)', ''), 
-          label_x = 0.10, label_y = 0.98,
+plot_grid(gradt, eke, labels = c('a)', 'b)'), 
+          label_x = 0.12, label_y = 0.98,
           label_fontface = "plain", 
           label_colour = "black",
-          rel_widths = c(1.75, 1))
+          rel_widths = c(1, 1))
 ```

### 2025-02-21 article/article.qmd
notes about grad temp

@@
 
-data <- purrr::map(Sys.glob(here("data/eke_mean/*hist*mean.nc")), function(f) {
+data <- purrr::map(c(Sys.glob(here("data/eke_mean/*hist*mean.nc")), Sys.glob(here("data/deltat/*hist*mean.nc"))), function(f) {
 
@@
   }
-  file_scenario <- here(paste0("data/eke_mean/", meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585_mean.nc"))
+  file_scenario <- Sys.glob(here(paste0("data/", c("eke_mean", "deltat"), "/", meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585_mean.nc")))
 
@@
 In this paper we build on the drops in temperature methodology to evaluate the change in frequency of cold fronts in a warmer future compared with the recent past using a set of models from CMIP6.
-In an effort to understand possible causes to the potential changes, we explored the relationship between changes in front frequency and transient eddy kinetic energy, that is directly related to storm tracks.
-Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@ogorman2010; @tamarin2017; @chemke2022], and these changes could potentially influence the location and intensity of future cold fronts.
+In an effort to understand possible causes to the potential changes, we first studied the response of the meridional temperature gradient and it relationship with the change in front frequency.
+We explored the relationship between changes in front frequency and transient eddy kinetic energy, that is directly related to storm tracks.
+Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@ogorman2010; @tamarin2017; @chemke2022], and these changes could also influence the location and intensity of future cold fronts.
 
@@
 
+### Meridional temperature gradient
+
+We calculated the meridional temperature gradient ($\partial T/ \partial y$) at each grid point using central finite differences for the temperature at 850 hPa.
+We then area-averaged $\partial T/ \partial y$ over different regions, i.e.
+between 25$^{\circ}$ and 65$^{\circ}$ S.
+This process produces the same results as if we calculate the temperature difference between2 latitudes but give us the advantage of calculating the mean $\partial T/ \partial y$ for regions over land and sea independently.
+
 ### Storm Tracks estimation
@@
 This pattern indicates that the areas of higher front frequency will shift poleward in the SSP5-8.5.
-
-**Increase over land vs decrease over sea could be related also to the thermal inertia / land get hotter?**
+This region is also associated with an increase in $\partial T/ \partial y$ in storm track activity (not show **or supplemental material**).
 
@@
 
-**why? **
-
 ```{r}
@@
 
-@fig-eke explores the relationship between the change in EKE and frequency of $\Delta T$ comparing the scenario SSP5-8.5 (2080-2100) and the recent past (1980-2000).
-In the Northern Hemisphere, both EKE and the frequency of $\Delta T$ show mainly negative changes (@fig-eke a).
-The zonal mean shows that this positive relationship holds well over oceans with negative changes of $\Delta T$ frequency and EKE around 40$^{\circ}$ N (@fig-eke b).
-In the Southern Hemisphere, the overall relationship appears negative, primarily due to increasing $\Delta T$ frequency over land while EKE maxima shifts poleward.
-**why?
-are we detecting changes in temp not related to cyclones and frontal things?** Over the sea, the changes are consistent with a poleward shift of the storm tracks (already reported by @ogorman2010, @tamarin2017, @chemke2022) and the maximum frequency of $\Delta T$.
+@fig-eke explores the relationship between the change in the frequency of $\Delta T$ and the response of $\partial T / \partial y $ and EKE by comparing the scenario SSP5-8.5 (2080-2100) and the recent past (1980-2000).
+There is a positive relationship between the changes in $\Delta T$ and $\partial T / \partial y$ in both hemispheres.
+The reduction in the frequency of fronts in the Northern Hemisphere could be associated to the decrease of $\partial T / \partial y$ in low levels already shown by other authors [e.g.
+@catto2014] while the opposite is true for the Southern Hemisphere (@fig-eke a).
+We see a similar relationship between changes in $\Delta T$ and EKE except for the Southern Hemisphere over land. The reason for this it is difference could be associated to many factors. 
+The change of the stormtracks in the SH is mainly associated to a poleward shift and may not have an effect on frequency of fronts over land the same way the $\partial T / \partial y$ does.
+@harvey2013 found that the stormtrack response in the SH during summer is associated to the change in the low level $\partial T / \partial y$ mainly bellow 65$^{\\circ}$ S, i.e.
+over the southern oceans.
 

### 2025-05-23 article/article.qmd
Add sst data, hoping it will be interesting!

@@
     email: martin.singh@monash.edu
+  - name: Michael J. Reeder
+    affiliations:
+      - name: Monash University
+        department: School of Earth, Atmosphere and Environment
+        region: Victoria
+        country: Australia
+    orcid: 0000-0001-8715-check
+    email: fake_email@fakeyfake.com
 abstract: |
@@
 
-data <- purrr::map(c(Sys.glob(here("data/eke_mean/*hist*mean.nc")), Sys.glob(here("data/deltat/*hist*mean.nc"))), function(f) {
+file_list <- c(Sys.glob(here("data/eke/*hist*mean.nc")), 
+               Sys.glob(here("data/deltat/*hist*mean.nc")),
+               Sys.glob(here("data/tos/*hist*mean.nc")))
+
+data <- purrr::map(file_list, function(f) {
 
   message(f)
-  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_{scenario}_mean.nc", "{var}_{model}_{scenario}_mean.nc"))
+  meta <- unglue::unglue(basename(f), "{var}_{model}_{scenario}_mean.nc")
 
@@
     var <- "ua"
+  } else if (meta[[1]][["var"]] == "deltap97.5") {
+    var <- "tasmax"
+    meta[[1]][["var"]] <- "deltat"
   } else {
-    var <- "tasmax"
+    var <- "tos"
   }
-  file_scenario <- Sys.glob(here(paste0("data/", c("eke_mean", "deltat"), "/", meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585_mean.nc")))
+
+  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/*_", meta[[1]][["model"]], "_ssp585_mean.nc")))
 
@@
 High topography areas are gray-shaded.
-Dotted areas show where at least 90% of the models agree in the sign of change"
+Dotted areas show where at least 90% of the models agree in the sign of change."
 
@@
 
-@fig-eke explores the relationship between the change in the frequency of $\Delta T$ and the response of $\partial T / \partial y $ and EKE by comparing the scenario SSP5-8.5 (2080-2100) and the recent past (1980-2000).
+@fig-eke explores the relationship between the change in the frequency of $\Delta T$ and the response of $\partial T / \partial y$ and EKE by comparing the scenario SSP5-8.5 (2080-2100) and the recent past (1980-2000).
 There is a positive relationship between the changes in $\Delta T$ and $\partial T / \partial y$ in both hemispheres.
@@
 The change of the stormtracks in the SH is mainly associated to a poleward shift and may not have an effect on frequency of fronts over land the same way the $\partial T / \partial y$ does.
-@harvey2013 found that the stormtrack response in the SH during summer is associated to the change in the low level $\partial T / \partial y$ mainly bellow 65$^{\\circ}$ S, i.e.
+@harvey2014 found that the stormtrack response in the SH during summer is associated to the change in the low level $\partial T / \partial y$ mainly bellow 65$^{\circ}$ S, i.e.
 over the southern oceans.

### 2025-05-26 article/article.qmd
Add SST gradient

@@
 file_list <- c(Sys.glob(here("data/eke/*hist*mean.nc")), 
-               Sys.glob(here("data/deltat/*hist*mean.nc")),
-               Sys.glob(here("data/tos/*hist*mean.nc")))
+               Sys.glob(here("data/deltat/*hist*mean.nc")))
 
-data <- purrr::map(file_list, function(f) {
+delta <- purrr::map(file_list, function(f) {
 
@@
 
-tas <- purrr::map(Sys.glob(here("data/tas/*hist*.nc")), function(f) {
+tas <- purrr::map(Sys.glob(here("data/tas/*hist*mean.nc")), function(f) {
 
   message(f)
-  meta <- unglue::unglue(basename(f), patterns = c("{var}_day_{model}_{scenario}_{member}_{grid}_20150101-21001231.nc"))
+  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_historical_mean.nc"))
 
   file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
-                                        meta[[1]][["var"]], "_day_", meta[[1]][["model"]], "_ssp585*")))
+                                        meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585*")))
 
@@
     _[, let(model = meta[[1]][["model"]],
-            var = "dt_dy",
+            var = var,
             lat = round(lat, digits = 8),
@@
 
-data <- rbind(data, tas) |> 
-  _[, let(land = MaskLand(lon, lat))] |> 
-  _[, let(land = factor(land, labels = c("Sea", "Land")))]
+tos <- purrr::map(Sys.glob(here("data/tos/*hist*mean.nc")), function(f) {
+
+  message(f)
+  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_historical_mean.nc"))
+
+  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
+                                        meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585*")))
+
+  var <- meta[[1]][["var"]]
+
+  ReadNetCDF(f, vars = c(hist = var)) |> 
+    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
+    _[, let(model = meta[[1]][["model"]],
+            lat = round(lat, digits = 8),
+            var = var,
+            dt_dy_hist = metR:::.derv(hist, lat)*(180/pi/6371000),
+            dt_dy_ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
+    _[, let(cos_lat = cos(lat*180/pi))] |> 
+    _[, let(dt_dx_hist = metR:::.derv(hist, lon)*(180/pi/6371000)*cos_lat,
+            dt_dx_ssp585 = metR:::.derv(ssp585, lon)*(180/pi/6371000)*cos_lat), by = .(lat)] |> 
+    _[, let(change = sqrt(dt_dx_ssp585^2 + dt_dy_ssp585^2) / sqrt(dt_dx_hist^2 + dt_dy_hist^2),
+            diff = sqrt(dt_dx_ssp585^2 + dt_dy_ssp585^2) - sqrt(dt_dx_hist^2 + dt_dy_hist^2))] |> 
+    _[, .(time, lat, lon, hist, ssp585, model, var, change, diff)]
+
+}) |> rbindlist()
+
+mask <- tos[var == "tos" & model == "NorESM2-MM", .(lon, lat, hist)] |> 
+  _[, let(land = fifelse(!is.na(hist), "Land", "Sea"))] |> 
+  _[, .(lon, lat, land)]
+
+data <- rbindlist(list(delta, tas, tos)) |> 
+  mask[x = _, on = .NATURAL]
+  # _[, let(land = MaskLand(lon, lat))] |> 
+  # _[, let(land = factor(land, labels = c("Sea", "Land")))]
 

### 2025-06-19 article/article.qmd
Small changes to text and figures after reading the paper for the millonsth time

@@
 keypoints: 
-  - Day-to-day drops in maximun temperature are identified as a proxy for cold fronts.  
-  - SSP5-8.5 from CMIP show a decrease in frequency of drops in temperature on the NH and a poleward shift over the oceans in the SH
-  - storm tracks?
+  - Day-to-day drops in maximun temperature are used as a proxy for cold fronts.  
+  - SSP5-8.5 CMIP6 scenario show a decrease in frequency of drops in temperature on the NH and a poleward shift over the oceans in the SH and an increase over land.
+  - Relationship with temperature gradient?
 bibliography: cold-fronts.bib  
@@
 file_list <- c(Sys.glob(here("data/eke/*hist*mean.nc")),
-               Sys.glob(here("data/deltat/*hist*mean.nc")))
+               Sys.glob(here("data/deltat/*hist*mean.nc")),
+               Sys.glob(here("data/dQ-dy/*hist*mean.nc")))
 
@@
   } else {
-    var <- "tos"
+    var <- "hfss"
   }
@@
             dt_dy_ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
-    _[, let(cos_lat = cos(lat*180/pi))] |> 
+    _[, let(cos_lat = cos(lat*pi/180))] |> 
     _[, let(dt_dx_hist = metR:::.derv(hist, lon)*(180/pi/6371000)*cos_lat,
@@
 Identifying fronts can be a complex task.
-To date, several computational methods have been developed that use different characteristics of fronts.
+To date, several computational methods have been developed using different characteristics of fronts.
 Some of these methods look for thermodynamic changes associated with temperature and humidity in different atmospheric levels [@berry2011], others identify changes in the wind field [@simmonds2012].
@@
 
-For this study, we use day-to-day changes in near-surface temperature (or drops in temperature) to estimate the passage of fronts [@reeder2015].
+For this study, we use day-to-day changes in near-surface maximun temperature (or drops in temperature) to estimate the passage of fronts [@reeder2015].
 This very simple method can be applied to both data from station observations and gridded data (with varying resolution, up to 2.5 degrees [@king2024]).
@@
 In an effort to understand possible causes to the potential changes, we first studied the response of the meridional temperature gradient and it relationship with the change in front frequency.
-We explored the relationship between changes in front frequency and transient eddy kinetic energy, that is directly related to storm tracks.
+We explored the relationship between changes in front frequency and transient eddy kinetic energy, as is directly related to storm tracks.
 Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@ogorman2010; @tamarin2017; @chemke2022], and these changes could also influence the location and intensity of future cold fronts.
@@
 We selected these models based on the data available for the  historical and scenario experiments stored on the Australian National Computational Infrastructure (NCI) replica data store.
-
-While all models have a nominal resolution of 100 km, we bilinear interpolated them onto the same grid to facilitate comparison.
 
@@
 We then area-averaged $\partial T/ \partial y$ over different regions, i.e.
-between 25$^{\circ}$ and 65$^{\circ}$ S.
+between 25$^{\circ}$ and 65$^{\circ}$ latitude for each hemisphere over land and sea.
 This process produces the same results as if we calculate the temperature difference between2 latitudes but give us the advantage of calculating the mean $\partial T/ \partial y$ for regions over land and sea independently.
@@
 
-To estimate the storm track location we use the transient eddy kinetic energy (EKE), used by many others to estimate storm tracks [@ogorman2010; @chemke2022].
+To estimate the storm track location we calculated the transient eddy kinetic energy (EKE), used by many others to estimate storm tracks [@ogorman2010; @chemke2022].
 We calculated EKE using a Fast Fourier Transform high-pass filter applied to daily horizontal wind at 300 hPa, with a 6-day cutoff for each model.
@@
 
-@fig-trend shows the change in $\Delta T$ frequency comparing the last 20 years of the 21st century (SSP5-8.5) and
-the last 20 years of the 20th century (historical run) in a) and the zonal mean for each latitude over sea b) and over land c).
+@fig-trend shows the change in $\Delta T$ frequency comparing the yearly averaged $\Delta T$ frequency for last 20 years of the 21st century (SSP5-8.5 experiment) and
+the last 20 years of the 20th century (historical experiment) in a) and the zonal mean of $\Delta T$ frequency for each latitude over sea b) and over land c).
 In the Northern Hemisphere over land, there is a general decrease in temperature drops, particularly north of 40$^{\circ}$ latitude, except in western and central Europe where there is an increase.
-This result is also observed by @catto2014 using CMIP5 and a different methodology to identify fronts reinforcing the reliability of the methodology used here.
+This result is also observed by @catto2014 using CMIP5 and a different methodology to identify fronts reinforcing the reliability of the methodology used in this analysis.
 Over the oceans, there is also a decrease in the frequency of fronts that could be associated to a decrease in the meridional temperature gradient.
-The local maxima over the North Atlantic and negative values south to that area could indicate that the maximum frequency of $\Delta T$ is shifting toward the pole.
+The local maxima over the North Atlantic and negative values south to that area could indicate that the maximum frequency of $\Delta T$ is shifting toward the pole (**do we need to show or add a climatology to supplemental material for this analysis?**).
 
@@
 sig <- data |> 
-  _[(model %in% models_in_gadi) & var == "deltap97.5"] |>
+  _[(model %in% models_in_gadi) & var == "deltat"] |>
   dcast(lat + lon + model ~ var, value.var = "diff") |> 
-  _[, .(sig = sum(sign(deltap97.5), na.rm = TRUE)), by = .(lon, lat)]
+  _[, .(sig = sum(sign(deltat), na.rm = TRUE)), by = .(lon, lat)]
 
@@
   dcast(lat + lon + model ~ var, value.var = "diff") |> 
-  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lon, lat)] |> 
+  _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lon, lat)] |> 
   ggplot(aes(ConvertLongitude(lon), lat)) +
@@
 mean_trend <- data |> 
-  _[(model %in% models_in_gadi) & var == "deltap97.5"] |>
+  _[(model %in% models_in_gadi) & var == "deltat"] |>
   dcast(lat + lon + model ~ var, value.var = "diff") |> 
@@
   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land, model)] |> 
+  _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lat, land, model)] |> 
   ggplot(aes(mean_dt, lat)) +
@@
         panel.background = element_rect(fill = "#fbfbfb", color = NA),
+        panel.grid.minor = element_line(linewidth = 0.1),
         strip.text.x = element_text(size = 12),
@@
 plot_grid(map_trend, mean_trend, labels = c('a)', ''), 
-          label_x = 0, label_y = 0.9,
+          label_x = 0, label_y = 0.94,
           label_fontface = "plain", 
@@
 
-The mean time series of $\Delta T$ frequency in @fig-serie shows trends consistent with @fig-trend a).
+The mean time series of $\Delta T$ frequency for different latitude bands in @fig-serie shows trends consistent with @fig-trend a).
 Interestingly, the trend around 50$^{\circ}$ N in the Northern Hemisphere is negative over sea and land since the beginning of the 20th century.
-In the Southern Hemisphere over the oceans, the time series for 30$^{\circ}$ and 40$^{\circ}$ S show a decrease of drops in temperature and the opposite happens on the 50$^{\circ}$ latitude band, indicating again a possible poleward shift of cold fronts in the future.
+In the Southern Hemisphere over the oceans, the time series for 30$^{\circ}$ and 40$^{\circ}$ S latitude bands show a decrease in drops of temperature and the opposite happens on the 50$^{\circ}$ latitude band, indicating again a possible poleward shift of cold fronts in the future.
 It is important to note the variability between models in the subfigure next to each panel.
 Each box plot is calculated using the year 2080-2100 for each model and region.
-The 50$^{\circ}$ latitude band in both hemispheres show the bigger changes but also the highest variability between models, however at least 70% of the models agree on the sign of the change.
+The 50$^{\circ}$ latitude band in both hemispheres show the bigger changes but also the highest variability between models, however at least 90% of the models agree on the sign of the change.
 
@@
 #| fig-height: 5
-#| fig.cap: "Multimodel mean frequency of $\\Delta T$ for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different latitude bands.
-Left to each main figure the intermodel distribution is shown for the period 2080-2100."
+#| fig.cap: "Multimodel mean frequency of $\\Delta T$ for the historical period (1850-2014) and SSP5-8.5 (2015-2100) over land and sea for different 2° latitude bands centered at 30, 40 and 50°.
+Left to each main subfigure the intermodel distribution is shown for the period 2080-2100."
 
@@
                         lat %between% c(-30, -28), "30 - SH"))] |>
-  _[!is.na(serie) & var == "deltap97.5" & model %in% models_in_gadi] |>
+  _[!is.na(serie) & var == "deltat" & model %in% models_in_gadi] |>
   _[, .(mean_freq = weighted.mean(ssp585, cos(lat*pi/180))), by = .(serie, land, model)] |> 
@@
   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-  _[, let(serie = fcase(lat %between% c(48, 50), "50 - NH",
-                        lat %between% c(38, 40), "40 - NH",
-                        lat %between% c(28, 30), "30 - NH", 
-                        lat %between% c(-50, -48), "50 - SH",
-                        lat %between% c(-40, -38), "40 - SH",
-                        lat %between% c(-30, -28), "30 - SH"))] |>
+  _[, let(serie = fcase(lat %between% c(48, 50), "50° - NH",
+                        lat %between% c(38, 40), "40° - NH",
+                        lat %between% c(28, 30), "30° - NH", 
+                        lat %between% c(-50, -48), "50° - SH",
+                        lat %between% c(-40, -38), "40° - SH",
+                        lat %between% c(-30, -28), "30° - SH"))] |>
   _[!is.na(serie)] |> 
   _[, .(mean_freq = weighted.mean(mean_freq, cos(lat*pi/180))), by = .(land, serie, year, scenario)] |> 
-  _[!(land == "Land" & serie == "50 - SH")]
+  _[!(land == "Land" & serie == "50° - SH")]
 
@@
 ```
-To explore the distribution of the change, @fig-distribution shows the change in frequency of $\Delta T$ using different percentiles as thresholds from 10 to 0.1th.
-The percentile values are also calculated from the historical period between 1979 and 2014 for each grid point and model as in @fig-trend.
+To explore the distribution of the change, @fig-distribution shows the change in frequency of $\Delta T$ using different percentiles as thresholds from 10 to 0.1th to identify drops in temperature.
+The percentile values are again calculated from the historical period between 1979 and 2014 for each grid point and model as in @fig-trend.
 This figure shows that the change in the frequency of $\Delta T$ is bigger at lower (more extreme) percentile thresholds.
@@
 However, it is important to note that lower percentiles thresholds are associated to rarer events, meaning fewer occurrences per year.
-These results suggest that regions projected to experience more temperature drops in the future, such as land areas in midlatitudes in the Southern Hemisphere, will also experience more intense $\Delta T$, whereas regions with a decreasing frequency of events will see less intense drops.
+These results suggest that regions projected to experience more temperature drops in the future, such as land areas in midlatitudes in the Southern Hemisphere, will also experience more intense $\Delta T$, whereas regions with a decreasing frequency of events will see less intense drops in temperature.
 
@@
 
-@fig-eke explores the relationship between the change in the frequency of $\Delta T$ and the response of $\partial T / \partial y$ and EKE by comparing the scenario SSP5-8.5 (2080-2100) and the recent past (1980-2000).
+@fig-eke explores the relationship between the change in the frequency of $\Delta T$ and the response of $\partial T / \partial y$ and EKE to the response to the increse of atmospheric $CO_2$ by comparing the scenario SSP5-8.5 (2080-2100) and the recent past (1980-2000).
 There is a positive relationship between the changes in $\Delta T$ and $\partial T / \partial y$ in both hemispheres.
-The reduction in the frequency of fronts in the Northern Hemisphere could be associated to the decrease of $\partial T / \partial y$ in low levels already shown by other authors [e.g.
-@catto2014] while the opposite is true for the Southern Hemisphere (@fig-eke a).
-We see a similar relationship between changes in $\Delta T$ and EKE except for the Southern Hemisphere over land. The reason for this it is difference could be associated to many factors. 
+The reduction in the frequency of fronts in the Northern Hemisphere could be associated to the decrease of $\partial T / \partial y$ in low levels that has already been shown by other authors [e.g.
+@catto2014].
+The the opposite effect is present the Southern Hemisphere (@fig-eke a).
+We see a similar relationship between changes in $\Delta T$ and EKE indicating that the changes in the storm tracks affects the location and frecuency of cold fronts.
+The exception is the Southern Hemisphere over land,  the reason for this it is difference could be associated to many factors.
 The change of the stormtracks in the SH is mainly associated to a poleward shift and may not have an effect on frequency of fronts over land the same way the $\partial T / \partial y$ does.
-@harvey2014 found that the stormtrack response in the SH during summer is associated to the change in the low level $\partial T / \partial y$ mainly bellow 65$^{\circ}$ S, i.e.
+Supporting this, @harvey2014 found that the stormtrack response in the SH during summer is associated to the change in the low level $\partial T / \partial y$ mainly bellow 65$^{\circ}$ S, i.e.
 over the southern oceans.
@@
 #| fig-height: 4
-#| fig.cap: "Mean change in $\\Delta T$ frequency and a) $\\nabla _{y}$ and b) EKE for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model.
+#| fig.cap: "Mean change in $\\Delta T$ frequency and a) $\\partial T / \\partial y$ and b) EKE for SSP5-8.5 (2080-2100) compared with the historical period (1980-2000) for each model.
 Values are calculated for the northern (25$^{\\circ}$ to 65$^{\\circ}$) and southern (-25$^{\\circ}$ to -65$^{\\circ}$) hemispheres and over sea (circles) and land (triangles).
@@
   _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180), na.rm = TRUE),
-        mean_dt_dy = weighted.mean(dt_dy, w = cos(lat*pi/180), na.rm = TRUE),
-        mean_deltat = weighted.mean(deltap97.5, w = cos(lat*pi/180), na.rm = TRUE)), by = .(region, land, model)] |> 
+        mean_dt_dy = weighted.mean(tas, w = cos(lat*pi/180), na.rm = TRUE),
+        mean_deltat = weighted.mean(deltat, w = cos(lat*pi/180), na.rm = TRUE)), by = .(region, land, model)] |> 
   na.omit()
@@
   scale_color_manual(values = c("cyan4", "orange")) +
-  labs(x = latex2exp::TeX("$\\nabla _{y} T$ change (K/km)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
+  labs(x = latex2exp::TeX("$\\partial T / \\partial y$ change (K/km)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
        color = NULL, shape = NULL) +
@@
 #   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-#   _[, .(mean_dt = mean(deltap97.5, na.rm = TRUE)), by = .(lat, land)]
+#   _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lat, land)]
 # 

### 2025-06-19 article/article.qmd
Something broke, now plots make sense again

@@
 
+  # file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/*_", meta[[1]][["model"]], "_ssp585_mean.nc")))
+
   if (meta[[1]][["var"]] == "eke") {
@@
 
-tos <- purrr::map(Sys.glob(here("data/tos/*hist*mean.nc")), function(f) {
+# tos <- purrr::map(Sys.glob(here("data/tos/*hist*mean.nc")), function(f) {
+#   
+#   message(f)
+#   meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_historical_mean.nc"))
+#   
+#   file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
+#                                         meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585*")))
+#   
+#   var <- meta[[1]][["var"]]
+#   
+#   ReadNetCDF(f, vars = c(hist = var)) |> 
+#     _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
+#     _[, let(model = meta[[1]][["model"]],
+#             lat = round(lat, digits = 8),
+#             var = var,
+#             dt_dy_hist = metR:::.derv(hist, lat)*(180/pi/6371000),
+#             dt_dy_ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
+#     _[, let(cos_lat = cos(lat*pi/180))] |> 
+#     _[, let(dt_dx_hist = metR:::.derv(hist, lon)*(180/pi/6371000)*cos_lat,
+#             dt_dx_ssp585 = metR:::.derv(ssp585, lon)*(180/pi/6371000)*cos_lat), by = .(lat)] |> 
+#     _[, let(change = sqrt(dt_dx_ssp585^2 + dt_dy_ssp585^2) / sqrt(dt_dx_hist^2 + dt_dy_hist^2),
+#             diff = sqrt(dt_dx_ssp585^2 + dt_dy_ssp585^2) - sqrt(dt_dx_hist^2 + dt_dy_hist^2))] |> 
+#     _[, .(time, lat, lon, hist, ssp585, model, var, change, diff)]
+#   
+# }) |> rbindlist() 
+# 
+# mask <- tos[var == "tos" & model == "NorESM2-MM", .(lon, lat, hist)] |> 
+#   _[, let(land = fifelse(!is.na(hist), "Land", "Sea"))] |> 
+#   _[, .(lon, lat, land)]
 
-  message(f)
-  meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_historical_mean.nc"))
-
-  file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
-                                        meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585*")))
-
-  var <- meta[[1]][["var"]]
-
-  ReadNetCDF(f, vars = c(hist = var)) |> 
-    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
-    _[, let(model = meta[[1]][["model"]],
-            lat = round(lat, digits = 8),
-            var = var,
-            dt_dy_hist = metR:::.derv(hist, lat)*(180/pi/6371000),
-            dt_dy_ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
-    _[, let(cos_lat = cos(lat*pi/180))] |> 
-    _[, let(dt_dx_hist = metR:::.derv(hist, lon)*(180/pi/6371000)*cos_lat,
-            dt_dx_ssp585 = metR:::.derv(ssp585, lon)*(180/pi/6371000)*cos_lat), by = .(lat)] |> 
-    _[, let(change = sqrt(dt_dx_ssp585^2 + dt_dy_ssp585^2) / sqrt(dt_dx_hist^2 + dt_dy_hist^2),
-            diff = sqrt(dt_dx_ssp585^2 + dt_dy_ssp585^2) - sqrt(dt_dx_hist^2 + dt_dy_hist^2))] |> 
-    _[, .(time, lat, lon, hist, ssp585, model, var, change, diff)]
-
-}) |> rbindlist()
-
-mask <- tos[var == "tos" & model == "NorESM2-MM", .(lon, lat, hist)] |> 
-  _[, let(land = fifelse(!is.na(hist), "Land", "Sea"))] |> 
-  _[, .(lon, lat, land)]
-
-data <- rbindlist(list(delta, tas, tos)) |> 
-  mask[x = _, on = .NATURAL]
-  # _[, let(land = MaskLand(lon, lat))] |> 
-  # _[, let(land = factor(land, labels = c("Sea", "Land")))]
+data <- rbindlist(list(delta, tas)) |> 
+  # mask[x = _, on = .NATURAL]
+  _[, let(land = MaskLand(lon, lat))] |>
+  _[, let(land = factor(land, labels = c("Sea", "Land")))]
 
@@
 model_spread <- data |> 
-  _[, let(serie = fcase(lat %between% c(48, 50), "50 - NH",
-                        lat %between% c(38, 40), "40 - NH",
-                        lat %between% c(28, 30), "30 - NH", 
-                        lat %between% c(-50, -48), "50 - SH",
-                        lat %between% c(-40, -38), "40 - SH",
-                        lat %between% c(-30, -28), "30 - SH"))] |>
-  _[!is.na(serie) & var == "deltat" & model %in% models_in_gadi] |>
+  _[var == "deltat" & model %in% models_in_gadi] |> 
+  _[, let(serie = fcase(lat %between% c(48, 50), "50° - NH",
+                        lat %between% c(38, 40), "40° - NH",
+                        lat %between% c(28, 30), "30° - NH", 
+                        lat %between% c(-50, -48), "50° - SH",
+                        lat %between% c(-40, -38), "40° - SH",
+                        lat %between% c(-30, -28), "30° - SH"))] |>
+  _[!is.na(serie)] |>
   _[, .(mean_freq = weighted.mean(ssp585, cos(lat*pi/180))), by = .(serie, land, model)] |> 
@@
   _[, let(scenario = "SSP5-8.5")] |> 
-  _[!(land == "Land" & hemisphere == "SH" & lat == 50)]
+  _[!(land == "Land" & hemisphere == "SH" & lat == "50°")]
 
@@
 dist |> 
-  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50 - NH",
-                        lat %between% c(38, 40) & land == "Sea", "40 - NH",
-                        lat %between% c(28, 30) & land == "Sea", "30 - NH", 
-                        lat %between% c(48, 50) & land == "Land", "50 - NH",
-                        lat %between% c(38, 40) & land == "Land", "40 - NH",
-                        lat %between% c(28, 30) & land == "Land", "30 - NH",
-                        lat %between% c(-50, -48) & land == "Land", "50 - SH",
-                        lat %between% c(-40, -38) & land == "Land", "40 - SH",
-                        lat %between% c(-30, -28) & land == "Land", "30 - SH",
-                        lat %between% c(-50, -48) & land == "Sea", "50 - SH",
-                        lat %between% c(-40, -38) & land == "Sea", "40 - SH",
-                        lat %between% c(-30, -28) & land == "Sea", "30 - SH"))] |>
+  _[, let(serie = fcase(lat %between% c(48, 50) & land == "Sea", "50° - NH",
+                        lat %between% c(38, 40) & land == "Sea", "40° - NH",
+                        lat %between% c(28, 30) & land == "Sea", "30° - NH", 
+                        lat %between% c(48, 50) & land == "Land", "50° - NH",
+                        lat %between% c(38, 40) & land == "Land", "40° - NH",
+                        lat %between% c(28, 30) & land == "Land", "30° - NH",
+                        lat %between% c(-50, -48) & land == "Land", "50° - SH",
+                        lat %between% c(-40, -38) & land == "Land", "40° - SH",
+                        lat %between% c(-30, -28) & land == "Land", "30° - SH",
+                        lat %between% c(-50, -48) & land == "Sea", "50° - SH",
+                        lat %between% c(-40, -38) & land == "Sea", "40° - SH",
+                        lat %between% c(-30, -28) & land == "Sea", "30° - SH"))] |>
   _[!is.na(serie)] |> 
@@
   dcast(land + serie + percentile ~ scenario, value.var = "mean_freq") |> 
+  _[, let(historical = fifelse(serie  == "50° - SH" & land == "Land", 0, historical),
+          ssp585 = fifelse(serie  == "50° - SH" & land == "Land", 0, ssp585))] |> 
   tidyr::separate(serie, into = c("lat", "hemisphere"), sep = " - ") |> 

### 2025-06-26 article/article.qmd
Ready for co-authors, maybe?

@@
 ---
-title: "Preparing your manuscript"
+title: "Future changes in cold front frequency and location in CMIP6 climate projections"
 format:
@@
 abstract: |
-  The abstract (1) states the nature of the investigation and (2) summarizes the important conclusions. The abstract should be suitable for indexing. Your abstract should:
-
-    - Be set as a single paragraph.
-    - Be less than 250 words for all journals except GRL, for which the limit is 150 words.
-    - Not include table or figure mentions.
-    - Avoid reference citations unless dependent on or directly related to another paper (e.g., companion, comment, reply, or commentary on another paper(s)).
-AGU’s Style Guide discusses formatting citations in abstracts.
-    - Define all abbreviations.
+  We use day-to-day drops in surface maximum temperature as a proxy to assess future changes in cold front frequency in a set of CMIP6 climate models under the SSP5-8.5 scenario.
+We find a general decrease in cold front frequency across the Northern Hemisphere, particularly over the oceans, and a poleward shift in the Southern Hemisphere, with increased frequency over midlatitude land regions.
+The frequency of strong temperature drops correlates positively with the meridional temperature gradient and transient eddy kinetic energy (EKE), highlighting the role of large-scale baroclinicity and storm track changes in future frontal behavior.
+Over Southern Hemisphere land areas, this relationship weakens, suggesting the influence of regional thermodynamic processes.
 plain-language-summary: |
-  A Plain Language Summary (PLS) can be an incredibly effective science communication tool.
-By summarizing your paper in non-technical terms, you can explain your research and its relevance to a much broader audience.
-A PLS is required for submissions to AGU Advances, G-Cubed, GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets, JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of Geophysics, but optional for other journals.
-A PLS should be no longer than 200 words and should be free of jargon, acronyms, equations, and any technical information that would be unknown to people from outside your scientific discipline.
-Read our tips for creating an effective PLS.
+  Cold fronts, weather systems associated with strong temperature drops and sometimes rain and strong wind, play a key role in the day-to-day weather in midlatitudes.
+In this study, we use a simple method based on changes in maximun temperature from one day to the next to identify cold fronts in climate simulations.
+Looking at a future cliamte projection for ten global models, we find that cold fronts are expected to become less frequent in much of the Northern Hemisphere, especially over the oceans.
+In the Southern Hemisphere, however, cold fronts are projected to shift closer to the South Pole and may become more frequent over land areas like southern Australia and southern Africa.
+These changes are linked to shifts in the temperature contrast between the equator and the poles and the paths taken by storm systems.
 keywords: []
@@
   - SSP5-8.5 CMIP6 scenario show a decrease in frequency of drops in temperature on the NH and a poleward shift over the oceans in the SH and an increase over land.
-  - Relationship with temperature gradient?
-bibliography: cold-fronts.bib  
+  - Changes in cold front frequency coudl be related to changes in the meridional gradient of surface temperature.
+bibliography: [cold-fronts.bib, packages.bib]  
 citation:
@@
 library(data.table)
-library(lubridate)
+# library(lubridate)
 library(here)
@@
   meta <- unglue::unglue(basename(f), "{var}_{model}_{scenario}_mean.nc")
-
-  # file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/*_", meta[[1]][["model"]], "_ssp585_mean.nc")))
 
@@
 
-tas <- purrr::map(Sys.glob(here("data/tas/*hist*mean.nc")), function(f) {
+ta <- purrr::map(Sys.glob(here("data/ta/*hist*mean.nc")), function(f) {
 
@@
 
-  ReadNetCDF(f, vars = c(hist = var)) |> 
-    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
+  ReadNetCDF(f, vars = c(hist = var), subset = list(plev = 85000)) |> 
+    _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), 
+                                subset = list(plev = 85000),
+                                out = "vector")[[1]])] |> 
     _[, let(model = meta[[1]][["model"]],
@@
     _[, let(change = ssp585 / hist,
-            diff = ssp585 - hist)]
+            diff = ssp585 - hist,
+            plev = NULL)]
 
@@
 
-# tos <- purrr::map(Sys.glob(here("data/tos/*hist*mean.nc")), function(f) {
-#   
-#   message(f)
-#   meta <- unglue::unglue(basename(f), patterns = c("{var}_{model}_historical_mean.nc"))
-#   
-#   file_scenario <- Sys.glob(here(paste0("data/", meta[[1]][["var"]], "/", 
-#                                         meta[[1]][["var"]], "_", meta[[1]][["model"]], "_ssp585*")))
-#   
-#   var <- meta[[1]][["var"]]
-#   
-#   ReadNetCDF(f, vars = c(hist = var)) |> 
-#     _[, let(ssp585 = ReadNetCDF(file_scenario, vars = c(ssp585 = var), out = "vector")[[1]])] |> 
-#     _[, let(model = meta[[1]][["model"]],
-#             lat = round(lat, digits = 8),
-#             var = var,
-#             dt_dy_hist = metR:::.derv(hist, lat)*(180/pi/6371000),
-#             dt_dy_ssp585 = metR:::.derv(ssp585, lat)*(180/pi/6371000)), by = .(lon)] |> 
-#     _[, let(cos_lat = cos(lat*pi/180))] |> 
-#     _[, let(dt_dx_hist = metR:::.derv(hist, lon)*(180/pi/6371000)*cos_lat,
-#             dt_dx_ssp585 = metR:::.derv(ssp585, lon)*(180/pi/6371000)*cos_lat), by = .(lat)] |> 
-#     _[, let(change = sqrt(dt_dx_ssp585^2 + dt_dy_ssp585^2) / sqrt(dt_dx_hist^2 + dt_dy_hist^2),
-#             diff = sqrt(dt_dx_ssp585^2 + dt_dy_ssp585^2) - sqrt(dt_dx_hist^2 + dt_dy_hist^2))] |> 
-#     _[, .(time, lat, lon, hist, ssp585, model, var, change, diff)]
-#   
-# }) |> rbindlist() 
-# 
-# mask <- tos[var == "tos" & model == "NorESM2-MM", .(lon, lat, hist)] |> 
-#   _[, let(land = fifelse(!is.na(hist), "Land", "Sea"))] |> 
-#   _[, .(lon, lat, land)]
-
-data <- rbindlist(list(delta, tas)) |> 
-  # mask[x = _, on = .NATURAL]
+data <- rbindlist(list(delta, ta)) |> 
   _[, let(land = MaskLand(lon, lat))] |>
@@
 
-Cold fronts are an important feature of weather in extratropical regions and are partly responsible for the precipitation in these areas [@konstali2024].
-Cold fronts are also linked to extreme events, such as heavy precipitation [@catto2013] and severe wildfires in southern Australia [@cai2022].
+Cold fronts are an important feature of weather in extratropical regions and are partly responsible for the precipitation in these areas [@Konstali2024].
+Cold fronts are also linked to extreme events, such as heavy precipitation [@Catto2013] and severe wildfires in southern Australia [@Cai2022].
 
@@
 To date, several computational methods have been developed using different characteristics of fronts.
-Some of these methods look for thermodynamic changes associated with temperature and humidity in different atmospheric levels [@berry2011], others identify changes in the wind field [@simmonds2012].
+Some of these methods look for thermodynamic changes associated with temperature and humidity in different atmospheric levels [@Berry2011], others identify changes in the wind field [@Simmonds2012].
 These methods rely on the data being in a regular grid (e.g.
@@
 
-For this study, we use day-to-day changes in near-surface maximun temperature (or drops in temperature) to estimate the passage of fronts [@reeder2015].
-This very simple method can be applied to both data from station observations and gridded data (with varying resolution, up to 2.5 degrees [@king2024]).
-Previous studies have shown that this method is suitable for the detection of strong cold fronts in southeastern Australia [@cai2022; @reeder2015].
-Recently, @king2024 demonstrated that this method can successfully identify cold fronts with results comparable to those obtained with other methodologies.
-Furthermore, @king2024 calculated a climatology of front frequency using global climate models from phase 6 of the Coupled Model Intercomparison Project (CMIP6).
+For this study, we use day-to-day changes in near-surface maximun temperature (or drops in temperature) to estimate the passage of fronts [@Reeder2015].
+This very simple method can be applied to both data from station observations and gridded data (with varying resolution, up to 2.5 degrees [@King2024]).
+Previous studies have shown that this method is suitable for the detection of strong cold fronts in southeastern Australia [@Cai2022; @Reeder2015].
+Recently, @King2024 demonstrated that this method can successfully identify cold fronts with results comparable to those obtained with other methodologies.
+Furthermore, @King2024 calculated a climatology of front frequency using global climate models from phase 6 of the Coupled Model Intercomparison Project (CMIP6).
 While the models tend to overestimate front frequency in polar regions and underestimate the occurrence of large temperature drops over midlatitude land regions compared with ERA5, they accurately represent the patterns of front frequency equatorward of 60°.
@@
 As global temperatures continue to rise, understanding how this warming will impact the occurrence of cold fronts is very important.
-A few studies have addressed this question, for example, @catto2014 used objective front identification [@berry2011] in 18 CMIP5 models and compared the changes between historical simulations between 1980-2005 and the Representative Concentration Pathway 8.5 between 2080-2100.
+A few studies have addressed this question, for example, @Catto2014 used objective front identification [@Berry2011] in 18 CMIP5 models and compared the changes between historical simulations between 1980-2005 and the Representative Concentration Pathway 8.5 between 2080-2100.
 They found an overall decrease in front frequency in the Northern Hemisphere, with a poleward shift in the maximum frequency and a significant decrease at high latitudes.
 In contrast, the Southern Hemisphere showed a poleward shift in the frequency maximum.
-On the other hand, @konstali2024, identified fronts in the CESM2-LE model simulations using the gradient of equivalent potential temperature in 850 hPa and found an increase in the frequency of fronts that could be related to fronts becoming larger or more elongated in both hemispheres.
+On the other hand, @Konstali2024, identified fronts in the CESM2-LE model simulations using the gradient of equivalent potential temperature in 850 hPa and found an increase in the frequency of fronts that could be related to fronts becoming larger or more elongated in both hemispheres.
 
@@
 We explored the relationship between changes in front frequency and transient eddy kinetic energy, as is directly related to storm tracks.
-Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@ogorman2010; @tamarin2017; @chemke2022], and these changes could also influence the location and intensity of future cold fronts.
+Several studies have shown a poleward shift on the storm tracks in future climate change scenarios, particularly in the Southern Hemisphere [@OGorman2010; @Tamarin2017; @Chemke2022], and these changes could also influence the location and intensity of future cold fronts.
 
@@
 
-For the analysis we use 10 models from phase 6 of the Coupled Model Intercomparison Project (CMIP6, @eyring2016), this includes Historical runs up to 2014 and the Shared Socioeconomic Pathways 5 (SSP5-8.5) from the Scenario Model Intercomparison Project (ScenarioMIP, @oneill2016) as described in @tbl-models.
+For the analysis we use 10 models from phase 6 of the Coupled Model Intercomparison Project (CMIP6, @Eyring2016), this includes Historical runs up to 2014 and the Shared Socioeconomic Pathways 5 (SSP5-8.5) from the Scenario Model Intercomparison Project (ScenarioMIP, @ONeill2016) as described in @tbl-models.
 We selected these models based on the data available for the  historical and scenario experiments stored on the Australian National Computational Infrastructure (NCI) replica data store.
@@
   _[order(source_id)] |> 
-  _[, let(references = c("@lovato2021; -@lovato2021a", "@consortiumec-earth2019; -@consortiumec-earth2019a",
-                         "@consortiumec-earth2020; -@consortiumec-earth2021", 
-                         "@consortiumec-earth2019b; -@consortiumec-earth2019c",
-                         "@consortiumec-earth2020a; -@consortiumec-earth2023",
-                         "@guo2018; -@guo2018a", "@volodin2019, @volodin2019a",
-                         "@volodin2019b; @volodin2019c", "@jungclaus2019; @schupfner2019",
-                         "@bentsen2019a; @bentsen2019b"))] |> 
+  _[, let(references = c("@Lovato2021; -@Lovato2021a", "@ConsortiumEC-Earth2019; -@ConsortiumEC-Earth2019a",
+                         "@ConsortiumEC-Earth2020; -@ConsortiumEC-Earth2021", 
+                         "@ConsortiumEC-Earth2019b; -@ConsortiumEC-Earth2019c",
+                         "@ConsortiumEC-Earth2020a; -@ConsortiumEC-Earth2023",
+                         "@Guo2018; -@Guo2018a", "@Volodin2019, @Volodin2019a",
+                         "@Volodin2019b; @Volodin2019c", "@Jungclaus2019; @Schupfner2019",
+                         "@Bentsen2019; @Bentsen2019a"))] |> 
   _[, let(source_id = paste0(source_id, "\n[", references, "]"))] |> 
@@
 The $\Delta T$ is calculated for all grid points and days over the studied periods.
-@king2024 defined a passage of a strong cold front as a $\Delta T \le -10^{\circ}C$, meaning where the maximum temperature on the following day is at least 10$^{\circ}$C colder.
-While this value is reasonable for detecting strong drops in temperature over land, it may underestimate drops over the ocean due to the smaller temperature variability **(standard deviation map for max temp in supplemental material?)**.
+@King2024 defined a passage of a strong cold front as a $\Delta T \le -10^{\circ}C$, meaning where the maximum temperature on the following day is at least 10$^{\circ}$C colder.
+While this value is reasonable for detecting strong drops in temperature over land, it may underestimate drops over the ocean due to the smaller temperature variability.
 For this reason, in this study we use the value of the percentile 2.5th (as we want to capture the most negative temperature drops) as a threshold for each grid point to identify drops in temperature, ie $\Delta T \le 2.5^{th}$ percentile.
-We chose this percentile threshold as the $\Delta T$ mean of values associated over land for midlatitudes (25$^{\circ}$ to 55$^{\circ}$) is close to $-10^{\circ}C$, the threshold used by @king2024.
+We chose this percentile threshold as the $\Delta T$ mean of values associated over land for midlatitudes (25$^{\circ}$ to 55$^{\circ}$) is close to $-10^{\circ}C$, the threshold used by @King2024.
 The percentiles are calculated over the historical period between 1979 and 2014 for each model at each grid point.
@@
 
-To estimate the storm track location we calculated the transient eddy kinetic energy (EKE), used by many others to estimate storm tracks [@ogorman2010; @chemke2022].
+To estimate the storm track location we calculated the transient eddy kinetic energy (EKE), used by many others to estimate storm tracks [@OGorman2010; @Chemke2022].
 We calculated EKE using a Fast Fourier Transform high-pass filter applied to daily horizontal wind at 300 hPa, with a 6-day cutoff for each model.
@@
 In the Northern Hemisphere over land, there is a general decrease in temperature drops, particularly north of 40$^{\circ}$ latitude, except in western and central Europe where there is an increase.
-This result is also observed by @catto2014 using CMIP5 and a different methodology to identify fronts reinforcing the reliability of the methodology used in this analysis.
+This result is also observed by @Catto2014 using CMIP5 and a different methodology to identify fronts reinforcing the reliability of the methodology used in this analysis.
 Over the oceans, there is also a decrease in the frequency of fronts that could be associated to a decrease in the meridional temperature gradient.
-The local maxima over the North Atlantic and negative values south to that area could indicate that the maximum frequency of $\Delta T$ is shifting toward the pole (**do we need to show or add a climatology to supplemental material for this analysis?**).
+The local maxima over the North Atlantic and negative values south to that area could indicate that the maximum frequency of $\Delta T$ is shifting toward the pole.
 
 In the Southern Hemisphere over land, the change is generally positive with 5 events per year in some areas.
-Other authors have also found an increase in the frequency of fronts in the south and east regions of Africa and Australia [@catto2014; @konstali2024].
+Other authors have also found an increase in the frequency of fronts in the south and east regions of Africa and Australia [@Catto2014; @Konstali2024].
 @fig-trend a and b) show over the southern oceans a decrease in the frequency of $\Delta T$ between 30$^{\circ}$ and 45$^{\circ}$ S (depending on the longitude) and an increase to the south.
 This pattern indicates that the areas of higher front frequency will shift poleward in the SSP5-8.5.
-This region is also associated with an increase in $\partial T/ \partial y$ in storm track activity (not show **or supplemental material**).
+This region is also associated with an increase in $\partial T/ \partial y$ in storm track activity (not shown).
 
@@
   geom_contour_fill(data = topo, aes(ConvertLongitude(lon), lat, z = topo), breaks = c(1300, Inf), fill = "grey90") +
-  coord_sf(expand = FALSE) +
+  coord_sf(expand = FALSE, xlim = c(-179, 180)) +
   labs(x = NULL, y = NULL, fill = "days/year") +
@@
   theme(legend.position = "inside",
-        legend.position.inside = c(0.78, 0.93),
+        legend.position.inside = c(0.78, 0.07),
         legend.direction = "horizontal",
@@
   theme(legend.position = "inside",
-        legend.position.inside = c(0.8, 0.93),
+        legend.position.inside = c(0.8, 0.07),
         legend.direction = "horizontal",
@@
 The reduction in the frequency of fronts in the Northern Hemisphere could be associated to the decrease of $\partial T / \partial y$ in low levels that has already been shown by other authors [e.g.
-@catto2014].
+@Catto2014].
 The the opposite effect is present the Southern Hemisphere (@fig-eke a).
@@
 The change of the stormtracks in the SH is mainly associated to a poleward shift and may not have an effect on frequency of fronts over land the same way the $\partial T / \partial y$ does.
-Supporting this, @harvey2014 found that the stormtrack response in the SH during summer is associated to the change in the low level $\partial T / \partial y$ mainly bellow 65$^{\circ}$ S, i.e.
+Supporting this, @Harvey2014 found that the stormtrack response in the SH during summer is associated to the change in the low level $\partial T / \partial y$ mainly bellow 65$^{\circ}$ S, i.e.
 over the southern oceans.
@@
   _[, .(mean_eke = weighted.mean(eke, w = cos(lat*pi/180), na.rm = TRUE),
-        mean_dt_dy = weighted.mean(tas, w = cos(lat*pi/180), na.rm = TRUE),
+        mean_dt_dy = weighted.mean(ta, w = cos(lat*pi/180), na.rm = TRUE),
         mean_deltat = weighted.mean(deltat, w = cos(lat*pi/180), na.rm = TRUE)), by = .(region, land, model)] |> 
+  _[, let(mean_dt_dy = fifelse(region == "NH", -mean_dt_dy, mean_dt_dy))] |> 
   na.omit()
@@
 gradt <- change |> 
-  _[, let(mean_dt_dy = fifelse(region == "NH", -mean_dt_dy, mean_dt_dy))] |> 
   ggplot(aes(mean_dt_dy*1000, mean_deltat)) +
@@
   scale_color_manual(values = c("cyan4", "orange")) +
-  labs(x = latex2exp::TeX("$\\partial T / \\partial y$ change (K/km)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
+  labs(x = latex2exp::TeX("$\\partial T / \\partial y$ change (K/km)"), 
+       y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
        color = NULL, shape = NULL) +
+  coord_cartesian(ylim = c(-4, 4)) +
   theme_minimal() +
   theme(legend.position = "inside",
-        legend.position.inside = c(0.83, 0.08),
+        legend.position.inside = c(0.73, 0.1),
         legend.direction = "horizontal",
-        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
-        legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
+        legend.box.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
+        legend.margin = margin(-2, 2, 1, 2, unit = "mm"),
         panel.background = element_rect(fill = "#fbfbfb", color = NA))
@@
   scale_color_manual(values = c("cyan4", "orange")) +
-  labs(x = latex2exp::TeX("EKE change ($m^2 s^{-2}$)"), y = latex2exp::TeX("$\\Delta T$ frequency change (days/year)"),
+  labs(x = latex2exp::TeX("EKE change ($m^2 s^{-2}$)"), 
+       y = NULL,
        color = NULL, shape = NULL) +
+  coord_cartesian(ylim = c(-4, 4)) +
   theme_minimal() +
-  theme(legend.position = "inside",
-        legend.position.inside = c(0.83, 0.08),
-        legend.direction = "horizontal",
-        legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
-        legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
+  theme(legend.position = "none",
+        # legend.position.inside = c(0.83, 0.08),
+        # legend.direction = "horizontal",
+        # legend.background = element_rect(fill = "#fbfbfb", color = "#fbfbfb"),
+        # legend.margin = margin(-3, 0, 0, 0, unit = "mm"),
         panel.background = element_rect(fill = "#fbfbfb", color = NA))
-
-# deltat_mean <- data |> 
-#   _[(model %in% models_in_gadi) & is.finite(change)] |>
-#   dcast(lat + lon + model ~ var, value.var = "diff") |> 
-#   _[, let(land = MaskLand(lon, lat))] |> 
-#   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-#   _[, .(mean_dt = mean(deltat, na.rm = TRUE)), by = .(lat, land)]
-# 
-# mean_trend_dt <- data |> 
-#   _[(model %in% models_in_gadi) & is.finite(change)] |>
-#   dcast(lat + lon + model ~ var, value.var = "diff") |> 
-#   _[, let(land = MaskLand(lon, lat))] |> 
-#   _[, let(land = factor(land, labels = c("Sea", "Land")))] |> 
-#   _[, .(mean_eke = mean(eke, na.rm = TRUE)), by = .(lat, land, model)] |> 
-#   ggplot(aes(mean_eke, lat)) +
-#   geom_hline(yintercept = 0, color = "grey80") +
-#   geom_vline(xintercept = 0, color = "grey80") +
-#   geom_path(aes(group = model), color = "grey70", alpha = 0.7) +
-#   stat_summary(orientation = "y", geom = "line", aes(color = "EKE")) +
-#   geom_path(data = deltat_mean, aes(x = mean_dt, color = "DeltaT")) +
-#   scale_color_manual(values = c("DeltaT" = "red4", "EKE" = "black"),
-#                      labels = c(latex2exp::TeX("$\\Delta T$"), "EKE")) +
-#   geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
-#             fill = "#fbfbfb",
-#             data = annotation,
-#             inherit.aes = FALSE) +
-#   annotate(geom = "rect", ymin = c(-90, -20, 90), ymax = c(-65, 20, 65), xmin = -Inf, xmax = Inf,
-#            fill = "grey93", alpha = 0.7) +
-#   scale_y_latitude(breaks = seq(-80, 80 , 20)) +
-#   facet_wrap(~land) +
-#   tagger::tag_facets(tag_pool = c("b", "c")) +
-#   labs(x = "Change", color = NULL) +
-#   theme_minimal() +
-#   theme(legend.position = "bottom",
-#         panel.background = element_rect(fill = "#fbfbfb", color = NA),
-#         strip.text.x = element_text(size = 12),
-#         plot.margin = unit(c(0, 0 , 0, -0.40), units = "cm"),
-#         tagger.panel.tag.text = element_text(size = 12))
 
@@
 
-What we did
+In this study, we use day-to-day drops in near-surface maximum temperature as a method to detect cold fronts and to evaluate changes in their frequency under future warming, based on a multi-model ensemble from CMIP6.
+By using a grid-point specific percentile threshold, we ensure a consistent detection of strong temperature drops across different regions, improving sensitivity particularly over oceans where the temperature variability is lower.
+However, is important to note that by using the 97.5th percentiles, we may filtering out week fronts and underestimating the overall frequency.
 
-Summary of results
+Our results show a general decrease in front frequency across the Northern Hemisphere, especially over oceans, and a poleward shift in frontal activity in the Southern Hemisphere.
+Over land in the Southern Hemisphere, notably in regions like southern Australia and Africa, we find an increase in front frequency and intensity.
+These findings align with previous studies using more complex frontal identification methods, strengthening confidence in our approach.
 
-Detection of fronts over oceans when using a constant threshold underestimates the fronts, using a percentile means detecting a similar number of fronts in each grid point.
-Weak fronts are not detected? 
-Contrast land - sea play a role here?
-Why EKE does not correlate well with deltaT in some areas? particularly SH-Land?
+We find a positive correlation between changes in $\Delta T$ frequency and the meridional temperature gradient.
+This suggests that baroclinicity remains a key control on frontogenesis in a warming climate.
+However, the expected correlation between front frequency and transient eddy kinetic energy (EKE) is weaker over Southern Hemisphere land regions.
+This may reflect the fact that storm track dynamics exert less direct control over frontal systems, which could be more influenced by mesoscale processes, land–sea contrast, and regional thermodynamic effects.
+
+One important factor not analyzed here is the role of sea surface temperature (SST) gradients in frontal development.
+[Perfit2017] showed that the surface sensible heat flux can amplify a fronts, through diabatic frontogenesis  in regions with strong SST gradients, such as the Gulf Stream or Kuroshio Current.
+However, [Reeder2021] showed that atmospheric fronts along SST fronts are amplified mainly by adiabatic frontogenesis, i.e.
+the deformation of the wind field that help increase the equivalent potencial temperature gradients.
+A projected weakening of SST gradients under climate change and in consequence, the sensible heat flux and wind circulation in the future could partially explain the changes in frequency of fronts along the west coast of North America and Asia.
+Further analysis is needed to quantify the contribution of these mechanisms to projected changes.
 
@@
 
-CLEX, NCI, etc
+This work was supported by part by the Australian Research Council Centre of Excellence for Climate Extremes grant CE170100023.
+This research was undertaken with the assistance of resources from the National Computational Infrastructure (NCI Australia), an NCRIS enabled capability supported by the Australian Government.
+P.C.
+also acknowledges the use of the following open-source software projects that supported data analysis, visualization, and reproducible reporting: R (v4.3.1, @R), metR [@metR], rcdo [@rcdo], data.table [@data.table], ggplot2 [@ggplot2], patchwork [@patchwork], cowplot [@cowplot], ggside [@ggside], Quarto [@quarto].
+
+**Should I also mention 21st century weather?**
 
